---
title: "02_Household_Forecast"
output:
  pdf_document: default
  html_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(knitr)
library(tidyverse)
library(forecast)
library(lubridate)
library(ggplot2)
library(tseries)
```

```{r source helpers}
source("../utils/one_step_model.R")
```

## Load Data

We use the train/test splits and differenced training data saved from the EDA notebook directly.

```{r load data}
train_df   <- readRDS("../output/train_df.rds")
test_df    <- readRDS("../output/test_df.rds")
macro_diff <- readRDS("../output/macro_df_diff.rds")

hh_train <- ts(train_df$hh_save, start = c(1977, 1), frequency = 4)
hh_test  <- ts(test_df$hh_save,  start = c(2015, 1), frequency = 4)
```

## ARIMA Order Identification

```{r ACF PACF of hh_save, fig.width=8, fig.height=6}

dhh_train <- diff(hh_train)

par(mfrow = c(2, 2))
acf(hh_train, lag.max = 20, main = "ACF: hh_save")
pacf(hh_train, lag.max = 20, main = "PACF: hh_save")
acf(dhh_train, lag.max = 20, main = "ACF: dhh")
pacf(dhh_train, lag.max = 20, main = "PACF: dhh")
par(mfrow = c(1, 1))
```

- The ACF of `hh_save` decays slowly, confirming non-stationarity and the need for differencing.
- After differencing, the ACF/PACF of `dhh` guides the choice of $p$ and $q$ for ARIMA.
- The candidate models we are considering are
  - ARIMA(1, 1, 1), ARIMA(0, 1, 1), ARIMA(2,1,0)
  - SARIMA(1, 1, 1)(1, 0, 0)[4], SARIMA(1, 1, 1)(0, 0, 1)[4], etc.

## One-Step Benchmark Models

All benchmarks forecast `hh_save` at level directly.

#### IID (Average Forecasts)

```{r IID}
rmse_tbl <- tibble(Model = character(), RMSE = double())
fc_store <- list()

# Average
res <- iid_fc(hh_train, hh_test)
rmse_tbl <- add_row(rmse_tbl, Model = "Average", RMSE = res$rmse)
fc_store$Average <- res$fc

print(res$rmse)
```

```{r IID by year}
# Average by Year
res <- iidbyyear_fc(hh_train, hh_test, periodicity = 4)
rmse_tbl <- add_row(rmse_tbl, Model = "Average by Year", RMSE = res$rmse)
fc_store$`Average by Year` <- res$fc
print(res$rmse)
```

```{r IID by period}
# Average by Period
res <- iidbyperiod_fc(hh_train, hh_test, periodicity = 4)
rmse_tbl <- add_row(rmse_tbl, Model = "Average by Period", RMSE = res$rmse)
fc_store$`Average by Period` <- res$fc
print(res$rmse)
```

### Persistence

```{r Persistence}
# Persistence forecast
res <- persist_fc(hh_train, hh_test)
rmse_tbl <- add_row(rmse_tbl, Model = "Persistence", RMSE = res$rmse)
fc_store$Persistence <- res$fc

print(res$rmse)
```

## Holt-Winters Variants

### SES

```{r Simple exponential smoothing}
# Simple Exponential Smoothing (SES)
hw_ses <- HoltWinters(hh_train, beta = FALSE, gamma = FALSE)
res <- esm_fc(hh_train, hh_test,
              alpha = hw_ses$alpha,
              level = as.numeric(hw_ses$coefficients[1]))
rmse_tbl <- add_row(rmse_tbl, Model = "SES", RMSE = res$rmse)
fc_store$SES <- res$fc

print(res$rmse)
```

### Holt Linear

```{r Holt linear}
hw_holt <- HoltWinters(hh_train, gamma = FALSE)
res <- lholt_fc(hh_train, hh_test,
                alpha = hw_holt$alpha,
                beta  = hw_holt$beta,
                level = as.numeric(hw_holt$coefficients[1]),
                slope = as.numeric(hw_holt$coefficients[2]))
rmse_tbl <- add_row(rmse_tbl, Model = "Holt Linear", RMSE = res$rmse)
fc_store$`Holt Linear` <- res$fc

print(res$rmse)
```

### Holt-Winters Seasonal Additive

```{r Holt-Winters additive}
hw_add <- HoltWinters(hh_train, seasonal = "additive")
res <- aseason_fc(hh_train, hh_test,
                  alpha  = hw_add$alpha,
                  beta   = hw_add$beta,
                  gamma  = hw_add$gamma,
                  level  = as.numeric(hw_add$coefficients[1]),
                  slope  = as.numeric(hw_add$coefficients[2]),
                  season = as.numeric(hw_add$coefficients[3:6]))
                  
rmse_tbl <- add_row(rmse_tbl, Model = "HW Additive", RMSE = res$rmse)
fc_store$`HW Additive` <- res$fc

print(res$rmse)
```

## ARIMA One-Step Forecast

```{r ARIMA one-step forecast}
# Safely extract ARIMA order
extract_order <- function(ord, keys, default = 0L) {
  vals <- ord[keys]
  vals[is.na(vals)] <- default
  as.integer(vals)
}

fit_arima <- auto.arima(hh_train,
                        seasonal = TRUE,
                        stepwise = FALSE,
                        approximation = FALSE)
# Print Auto ARIMA summary                   
summary(fit_arima)
```

```{r Alternative ARIMA specifications}
# Try alternative ARIMA specifications to check robustness of order selection
fit_arima_111 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  include.mean = TRUE
)

fit_arima_011 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  include.mean = TRUE
)

fit_arima_210 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(2, 1, 0),
  include.mean = TRUE
)

fit_arima_110 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 0),
  include.mean = TRUE
)

arima_tbl <- tibble(
  Model = c("ARIMA(1,1,1)", "ARIMA(0,1,1)", "ARIMA(2,1,0)", "ARIMA(1,1,0)"),
  AIC = round(c(fit_arima_111$fit$aic, fit_arima_011$fit$aic, 
                fit_arima_210$fit$aic, fit_arima_110$fit$aic), 2),
  RMSE = c(fit_arima_111$rmse, fit_arima_011$rmse, 
           fit_arima_210$rmse, fit_arima_110$rmse)
)

arima_tbl
```
```{r Add the best ARIMA}
arima_fcs <- list(
  "ARIMA(1,1,1)" = fit_arima_111$fc,
  "ARIMA(0,1,1)" = fit_arima_011$fc,
  "ARIMA(2,1,0)" = fit_arima_210$fc,
  "ARIMA(1,1,0)" = fit_arima_110$fc
)
arima_fits <- list(
  "ARIMA(1,1,1)" = fit_arima_111$fit,
  "ARIMA(0,1,1)" = fit_arima_011$fit,
  "ARIMA(2,1,0)" = fit_arima_210$fit,
  "ARIMA(1,1,0)" = fit_arima_110$fit
)
best_arima_idx   <- which.min(arima_tbl$RMSE)
best_arima_label <- arima_tbl$Model[best_arima_idx]
rmse_tbl <- add_row(rmse_tbl, Model = best_arima_label, RMSE = arima_tbl$RMSE[best_arima_idx])
fc_store[[best_arima_label]] <- arima_fcs[[best_arima_label]]
```

### ARIMA Residual Diagnostics

```{r ARIMA residual diagnostics, fig.width=10, fig.height=6}
best_arima_fit <- arima_fits[[best_arima_label]]
cat("Best ARIMA:", best_arima_label, "| AIC:", round(best_arima_fit$aic, 2), "\n")
forecast::checkresiduals(best_arima_fit)
```

The `checkresiduals()` function displays: (1) a time plot of residuals, (2) the ACF of residuals, (3) a histogram of residuals, and (4) a Ljung-Box test. If the residuals resemble white noise (no significant ACF spikes, roughly normal distribution, Ljung-Box p-value > 0.05), the ARIMA specification is adequate.

```{r Auto SARIMA}
ord <- forecast::arimaorder(fit_arima)
res <- sarima_one_step_fc(
  train    = hh_train,
  holdout  = hh_test,
  order    = extract_order(ord, c("p", "d", "q")),
  seasonal = extract_order(ord, c("P", "D", "Q")),
  include.mean = TRUE
)
rmse_tbl <- add_row(rmse_tbl, Model = "Auto ARIMA", RMSE = res$rmse)
fc_store$`Auto ARIMA` <- res$fc
print(res$rmse)
```

```{r Try alternative SARIMA specifications}
fit_sarima_111_100 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(1, 0, 0),
  include.mean = TRUE
)

fit_sarima_011_100 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(1, 0, 0),
  include.mean = TRUE
)

fit_sarima_111_101 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(1, 0, 1),
  include.mean = TRUE
)

fit_sarima_011_101 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(1, 0, 1),
  include.mean = TRUE
)

fit_sarima_111_001 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(0, 0, 1),
  include.mean = TRUE
)

fit_sarima_011_001 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(0, 0, 1),
  include.mean = TRUE
)

fit_sarima_111_002 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(0, 0, 2),
  include.mean = TRUE
)

fit_sarima_011_002 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(0, 0, 2),
  include.mean = TRUE
)

sarima_tbl <- tibble(
  Model = c(
    "SARIMA(1,1,1)(1,0,0)[4]",
    "SARIMA(0,1,1)(1,0,0)[4]",
    "SARIMA(1,1,1)(1,0,1)[4]",
    "SARIMA(0,1,1)(1,0,1)[4]",
    "SARIMA(1,1,1)(0,0,1)[4]",
    "SARIMA(0,1,1)(0,0,1)[4]",
    "SARIMA(1,1,1)(0,0,2)[4]",
    "SARIMA(0,1,1)(0,0,2)[4]"
  ),
  AIC = round(c(
    fit_sarima_111_100$fit$aic,
    fit_sarima_011_100$fit$aic,
    fit_sarima_111_101$fit$aic,
    fit_sarima_011_101$fit$aic,
    fit_sarima_111_001$fit$aic,
    fit_sarima_011_001$fit$aic,
    fit_sarima_111_002$fit$aic,
    fit_sarima_011_002$fit$aic
  ), 2),
  RMSE = c(
    fit_sarima_111_100$rmse,
    fit_sarima_011_100$rmse,
    fit_sarima_111_101$rmse,
    fit_sarima_011_101$rmse,
    fit_sarima_111_001$rmse,
    fit_sarima_011_001$rmse,
    fit_sarima_111_002$rmse,
    fit_sarima_011_002$rmse
  )
)

sarima_tbl
```

```{r Store the best SARIMA}
sarima_fcs <- list(
  "SARIMA(1,1,1)(1,0,0)[4]" = fit_sarima_111_100$fc,
  "SARIMA(0,1,1)(1,0,0)[4]" = fit_sarima_011_100$fc,
  "SARIMA(1,1,1)(1,0,1)[4]" = fit_sarima_111_101$fc,
  "SARIMA(0,1,1)(1,0,1)[4]" = fit_sarima_011_101$fc,
  "SARIMA(1,1,1)(0,0,1)[4]" = fit_sarima_111_001$fc,
  "SARIMA(0,1,1)(0,0,1)[4]" = fit_sarima_011_001$fc,
  "SARIMA(1,1,1)(0,0,2)[4]" = fit_sarima_111_002$fc,
  "SARIMA(0,1,1)(0,0,2)[4]" = fit_sarima_011_002$fc
)
sarima_fits <- list(
  "SARIMA(1,1,1)(1,0,0)[4]" = fit_sarima_111_100$fit,
  "SARIMA(0,1,1)(1,0,0)[4]" = fit_sarima_011_100$fit,
  "SARIMA(1,1,1)(1,0,1)[4]" = fit_sarima_111_101$fit,
  "SARIMA(0,1,1)(1,0,1)[4]" = fit_sarima_011_101$fit,
  "SARIMA(1,1,1)(0,0,1)[4]" = fit_sarima_111_001$fit,
  "SARIMA(0,1,1)(0,0,1)[4]" = fit_sarima_011_001$fit,
  "SARIMA(1,1,1)(0,0,2)[4]" = fit_sarima_111_002$fit,
  "SARIMA(0,1,1)(0,0,2)[4]" = fit_sarima_011_002$fit
)
best_sarima_idx   <- which.min(sarima_tbl$RMSE)
best_sarima_label <- sarima_tbl$Model[best_sarima_idx]
rmse_tbl <- add_row(rmse_tbl, Model = best_sarima_label, RMSE = sarima_tbl$RMSE[best_sarima_idx])
fc_store[[best_sarima_label]] <- sarima_fcs[[best_sarima_label]]
```

### SARIMA Residual Diagnostics

```{r SARIMA residual diagnostics, fig.width=10, fig.height=6}
best_sarima_fit <- sarima_fits[[best_sarima_label]]
cat("Best SARIMA:", best_sarima_label, "| AIC:", round(best_sarima_fit$aic, 2), "\n")
forecast::checkresiduals(best_sarima_fit)
```

## ARIMAX One-Step Forecast

### Exogenous Variable Selection via CCF

We use the differenced training data from the EDA to inspect cross-correlations with `dhh`. A variable is selected for ARIMAX only if it shows significant correlation at **negative lags** (the regressor leads `dhh`), to avoid overfitting with only 20 holdout observations.

```{r CCF variable selection, fig.width=10, fig.height=6}
candidates <- c("dcorp", "ldgdi", "ldcons", "dunemp", "dinterest", "ldcpi", "ldgpdi")

par(
  mfrow = c(2, 4),
  mar = c(5, 5, 3, 1),
  mgp = c(2, 0.8, 0),
  cex.main = 0.95,
  cex.lab = 0.9,
  cex.axis = 0.85
)
for (v in candidates) {
  ccf(na.omit(macro_diff[[v]]), na.omit(macro_diff$dhh),
      lag.max = 16,
      main = paste(v, "vs dhh"))
}
par(mfrow = c(1, 1))
```

```{r CCF values}
ccf_values <- sapply(candidates, function(v) {
  ccf_res <- ccf(na.omit(macro_diff[[v]]), na.omit(macro_diff$dhh), lag.max = 16, plot = FALSE)
  max_ccf <- max(abs(ccf_res$acf[ccf_res$lag < 0]))
  return(max_ccf)
})

# Identify the lag with the maximum absolute CCF for each candidate
ccf_lags <- sapply(candidates, function(v) {
  ccf_res <- ccf(na.omit(macro_diff[[v]]), na.omit(macro_diff$dhh), lag.max = 16, plot = FALSE)
  max_idx <- which.max(abs(ccf_res$acf[ccf_res$lag < 0]))
  return(ccf_res$lag[max_idx])
})

ccf_tbl <- tibble(
  Variable = candidates,
  Max_CCF = ccf_values,
  Lag_at_Max_CCF = ccf_lags
)

ccf_tbl
```

Based on the CCF evidence and the EDA findings, we select regressors with clear leading relationship:

- `ldgpdi`
- `ldcpi`

```{r ARIMAX preprocessing}
# Compute stationary exogenous variables across the full sample
arimax_df <- bind_rows(train_df, test_df) |>
  arrange(date) |>
  mutate(
    ldgpdi  = log(gpdi) - log(lag(gpdi)),
    ldcpi = log1p(cpi) - log1p(lag(cpi)),
    ldcpi_lag2 = lag(ldcpi, 2),
    ldgpdi_lag1 = lag(ldgpdi, 1),
    ldgpdi_lag3 = lag(ldgpdi, 3)
  ) |>
  drop_na(ldgpdi_lag1, ldcpi_lag2, ldgpdi_lag3) # Drop initial rows with NA due to differencing and lagging

arimax_vars <- c("ldgpdi_lag1", "ldcpi_lag2", "ldgpdi_lag3")

x_train <- arimax_df |> filter(date <= as.Date("2014-12-31")) |>
  select(all_of(arimax_vars)) |>
  as.matrix()
x_test <- arimax_df |> filter(date > as.Date("2014-12-31")) |>
  select(all_of(arimax_vars)) |>
  as.matrix()
```

### Fitting ARIMAX models

```{r ARIMAX one-step forecast}

# Align hh_save accordingly
hh_train_x <- ts(
  arimax_df |> filter(date <= as.Date("2014-12-31")) |> pull(hh_save),
  start = c(
    year(min(arimax_df$date)),
    quarter(min(arimax_df$date))
  ), frequency = 4
)

# Fit ARIMAX with selected regressors
fit_arimax <- auto.arima(hh_train_x,
                         xreg = x_train,
                         seasonal = FALSE,
                         stepwise = FALSE,
                         approximation = FALSE)
summary(fit_arimax)
```

```{r ARIMAX forecast}
arimax_213 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(2, 1, 3),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_212 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(2, 1, 2),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_112 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 2),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_111 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 1),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_011 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(0, 1, 1),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_tbl <- tibble(
  Model = c(
    "ARIMAX(2,1,3)",
    "ARIMAX(2,1,2)",
    "ARIMAX(1,1,2)",
    "ARIMAX(1,1,1)",
    "ARIMAX(0,1,1)"
  ),
  AIC = round(c(
    arimax_213$fit$aic,
    arimax_212$fit$aic,
    arimax_112$fit$aic,
    arimax_111$fit$aic,
    arimax_011$fit$aic
  ), 2),
  RMSE = c(
    arimax_213$rmse,
    arimax_212$rmse,
    arimax_112$rmse,
    arimax_111$rmse,
    arimax_011$rmse
  )
)

arimax_tbl
```

```{r Store the best ARIMAX}
arimax_fcs <- list(
  "ARIMAX(2,1,3)" = arimax_213$fc,
  "ARIMAX(2,1,2)" = arimax_212$fc,
  "ARIMAX(1,1,2)" = arimax_112$fc,
  "ARIMAX(1,1,1)" = arimax_111$fc,
  "ARIMAX(0,1,1)" = arimax_011$fc
)
arimax_fits <- list(
  "ARIMAX(2,1,3)" = arimax_213$fit,
  "ARIMAX(2,1,2)" = arimax_212$fit,
  "ARIMAX(1,1,2)" = arimax_112$fit,
  "ARIMAX(1,1,1)" = arimax_111$fit,
  "ARIMAX(0,1,1)" = arimax_011$fit
)
best_arimax_idx   <- which.min(arimax_tbl$RMSE)
best_arimax_label <- arimax_tbl$Model[best_arimax_idx]
rmse_tbl <- add_row(rmse_tbl, Model = best_arimax_label, RMSE = arimax_tbl$RMSE[best_arimax_idx])
fc_store[[best_arimax_label]] <- arimax_fcs[[best_arimax_label]]
```

### ARIMAX Residual Diagnostics

```{r ARIMAX residual diagnostics, fig.width=10, fig.height=6}
best_arimax_fit <- arimax_fits[[best_arimax_label]]
cat("Best ARIMAX:", best_arimax_label, "| AIC:", round(best_arimax_fit$aic, 2), "\n")
forecast::checkresiduals(best_arimax_fit)
```

## Theoretical Model (Precautionary Saving)

In Macroeconomics, the precautionary saving motive is modeled by the permanent income model. Where, assuming a representative household is constant relative risk averse, we have:
$$
C_t^{-\gamma} = \beta\mathbb{E}_t[(1+r_{t+1})C_{t+1}] 
$$
where $C_t$ is consumption, $\beta$ is the discount factor, $\gamma$ is the coefficient of relative risk aversion, and $r_{t+1}$ is the real interest rate. This implies that consumption growth depends on expected returns and the intertemporal elasticity of substitution.

Taking the log-transformation and Taylor expansion will eventually give us
$$
c_t = c_{t+1} - \frac{1}{\gamma} \ln \beta - \frac{1}{\gamma} r_{t+1}  - \frac{\gamma}{2} \text{Var}_t(\Delta c_{t+1}) - \varepsilon_{t+1}
$$
Where with the identity $s_t = y_t - c_t$ , the first difference of saving is captured by
$$
\Delta s_t = \Delta y_t - \Delta\mathbb{E}_t[c_{t+1}] + \frac{1}{\gamma}\Delta r_{t+1} + \frac{\gamma}{2} \Delta\text{Var}_t(\Delta c_{t+1}) + \varepsilon_{t+1}
$$
This suggests we can fit the following empirical model
$$
dhh_t = \alpha + \sum_{i=0}^p \phi_i dhh_{t-i} + \beta_1 dinterest_{t-1} + \beta_2 dunemp_{t-1} + \beta_3 ldgdi_{t-1} + \beta_4 ldgpdi_{t-1} + \varepsilon_t
$$
Where we use the unemployment and Gross Private Domestic Investment (GPDI) as proxies for uncertainty, and the real interest rate as the return variable. We lag all regressors by 1 quarter to avoid look-ahead bias in forecasting.

```{r Rainy day forecast}
pim_df <- bind_rows(train_df, test_df) |>
  arrange(date) |>
  mutate(
    dinterest = interest - lag(interest),
    dunemp = unemp - lag(unemp),
    ldgdi = log(gdi) - log(lag(gdi)),
    ldgpdi = log(gpdi) - log(lag(gpdi)),
    dinterest_lag1 = lag(dinterest, 1),
    dunemp_lag1 = lag(dunemp, 1),
    ldgdi_lag1 = lag(ldgdi, 1),
    ldgpdi_lag1 = lag(ldgpdi, 1)
  ) |>
  drop_na(ldgdi_lag1)

pim_train <- pim_df |> filter(date <= as.Date("2014-12-31"))|>
  select(dinterest_lag1, dunemp_lag1, ldgdi_lag1, ldgpdi_lag1)|>
  as.matrix()
pim_test <- pim_df |> filter(date > as.Date("2014-12-31"))|>
  select(dinterest_lag1, dunemp_lag1, ldgdi_lag1, ldgpdi_lag1)|>
  as.matrix()

hh_train_x <- ts(
  pim_df |> filter(date <= as.Date("2014-12-31"))|>pull(hh_save),
  start = c(
    year(min(pim_df$date)),
    quarter(min(pim_df$date))
  ), 
  frequency = 4
)
```

```{r Fit the rainy day model}
pim_fit <- auto.arima(hh_train_x, 
                      xreg = pim_train,
                      seasonal = FALSE,
                      stepwise = FALSE,
                      approximation = FALSE)
pim_fit
```
```{r Test alternatives}
pim_111 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 1),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)

pim_011 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(0, 1, 1),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)


pim_110 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 0),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)

pim_012 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(0, 1, 2),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)

pim_tbl <- tibble(
  Model = c(
    "PIM ARIMAX(1,1,1)",
    "PIM ARIMAX(0,1,1)",
    "PIM ARIMAX(1,1,0)",
    "PIM ARIMAX(0,1,2)"
  ),
  AIC = round(c(
    pim_111$fit$aic,
    pim_011$fit$aic,
    pim_110$fit$aic,
    pim_012$fit$aic
  ), 2),
  RMSE = c(
    pim_111$rmse,
    pim_011$rmse,
    pim_110$rmse,
    pim_012$rmse
  )
)

pim_tbl
```

```{r Store the best PIM ARIMAX}
pim_fcs <- list(
  "PIM ARIMAX(1,1,1)" = pim_111$fc,
  "PIM ARIMAX(0,1,1)" = pim_011$fc,
  "PIM ARIMAX(1,1,0)" = pim_110$fc,
  "PIM ARIMAX(0,1,2)" = pim_012$fc
)
pim_fits <- list(
  "PIM ARIMAX(1,1,1)" = pim_111$fit,
  "PIM ARIMAX(0,1,1)" = pim_011$fit,
  "PIM ARIMAX(1,1,0)" = pim_110$fit,
  "PIM ARIMAX(0,1,2)" = pim_012$fit
)
best_pim_idx   <- which.min(pim_tbl$RMSE)
best_pim_label <- pim_tbl$Model[best_pim_idx]
rmse_tbl <- add_row(rmse_tbl, Model = best_pim_label, RMSE = pim_tbl$RMSE[best_pim_idx])
fc_store[[best_pim_label]] <- pim_fcs[[best_pim_label]]
```

### PIM ARIMAX Residual Diagnostics

```{r PIM residual diagnostics, fig.width=10, fig.height=6}
best_pim_fit <- pim_fits[[best_pim_label]]
cat("Best PIM:", best_pim_label, "| AIC:", round(best_pim_fit$aic, 2), "\n")
forecast::checkresiduals(best_pim_fit)
```

## RMSE Ranking

```{r RMSE ranking}
rmse_tbl <- rmse_tbl |>
  mutate(RMSE = round(RMSE, 4)) |>
  arrange(RMSE)

rmse_tbl

best_model <- rmse_tbl$Model[1]
cat("Best model by one-step holdout RMSE:", best_model, "\n")
```

## Holdout Visualization

```{r holdout forecast plot, fig.width=10, fig.height=6}
fc_df <- tibble(date = test_df$date, Actual = as.numeric(hh_test))


for (m in names(fc_store)) {
  fc_df[[m]] <- as.numeric(fc_store[[m]])
}

fc_long <- fc_df |>
  pivot_longer(cols = -date, names_to = "Model", values_to = "value")|>
  filter(
    Model %in% c("Actual", "Average", "Persistence", 
                 "HW Additive", 
                 best_arima_label, 
                 best_sarima_label,
                 best_arimax_label,
                 best_pim_label)
  ) # Keep only benchmarks and the best HW and ARIMA/SARIMA models for clarity

ggplot(fc_long, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "One-step-ahead holdout forecasts of household net savings",
    x = "Quarter",
    y = "Household net savings",
    color = "Model",
    linetype = "Model"
  )
```

## Interpretation

- All benchmarks and ARIMA/ARIMAX models forecast `hh_save` directly; differencing is handled internally by `auto.arima` via the $d$ parameter.
- The ARIMAX uses only CCF-justified regressors (`r paste(arimax_vars, collapse = ", ")`) to avoid overfitting the small holdout window.
- The RMSE ranking table determines the best model for household net savings forecasting.

## Expanding Window Backtest

Our one-step-ahead evaluation above uses **fixed-parameter Kalman filtering**: estimate the model once on the training set (1977 Q1 – 2014 Q4), then run the Kalman filter through the holdout while keeping all parameters frozen. This is valid one-step-ahead forecasting — at each holdout time $t$, the forecast $\hat{y}_t$ is conditioned on $y_1, \dots, y_{t-1}$ — but parameter estimates are never updated.

An **expanding window backtest** re-estimates the model at every holdout step:

1. Fit model on $y_1, \dots, y_{T}$; forecast $y_{T+1}$.
2. Fit model on $y_1, \dots, y_{T+1}$; forecast $y_{T+2}$.
3. Continue until the end of the holdout.

**Is expanding window less biased?**

- **Yes, in general.** The fixed-parameter approach is equivalent to an expanding window that never updates parameters — it uses the *same* parameter estimates for all 20 holdout quarters. If the true DGP shifts over the holdout (e.g., structural break), stale parameters will systematically misforecast, biasing the RMSE upward. The expanding window adapts, yielding an RMSE that better reflects what a practitioner would achieve in real-time.
- **Trade-off:** Expanding window can overfit in small samples — each re-estimation uses slightly more data, but for early holdout quarters, the training set is essentially the same. For our 152-quarter training set and 20-quarter holdout, the effect is modest (training size changes from 152 to 171).
- **Computational cost:** Each model must be re-estimated $h$ times (here $h = 20$), so the expanding window is ~20× slower.

We compare the two approaches for the best ARIMA and SARIMA models below.

```{r Expanding window comparison, fig.width=10, fig.height=5}
# Parse the best ARIMA order from label
parse_arima_order <- function(label) {
  nums <- as.integer(regmatches(label, gregexpr("[0-9]+", label))[[1]])
  return(nums[1:3])
}

parse_sarima_orders <- function(label) {
  nums <- as.integer(regmatches(label, gregexpr("[0-9]+", label))[[1]])
  list(order = nums[1:3], seasonal = nums[4:6])
}

# Best ARIMA expanding window
best_arima_order <- parse_arima_order(best_arima_label)
ew_arima <- arima_expanding_window_fc(
  train = hh_train,
  holdout = hh_test,
  order = best_arima_order,
  include.mean = TRUE
)

# Best SARIMA expanding window
best_sarima_orders <- parse_sarima_orders(best_sarima_label)
ew_sarima <- sarima_expanding_window_fc(
  train = hh_train,
  holdout = hh_test,
  order = best_sarima_orders$order,
  seasonal = best_sarima_orders$seasonal,
  include.mean = TRUE
)

# Also Persistence expanding window = same as fixed (no parameters to update)
ew_persist <- persist_fc(hh_train, hh_test)

ew_comparison <- tibble(
  Model = c("Persistence",
            paste(best_arima_label, "(fixed)"),
            paste(best_arima_label, "(expanding)"),
            paste(best_sarima_label, "(fixed)"),
            paste(best_sarima_label, "(expanding)")),
  RMSE = round(c(ew_persist$rmse,
                  fc_store[[best_arima_label]] |> 
                    {\(fc) sqrt(mean((as.numeric(hh_test) - fc)^2))}(),
                  ew_arima$rmse,
                  fc_store[[best_sarima_label]] |> 
                    {\(fc) sqrt(mean((as.numeric(hh_test) - fc)^2))}(),
                  ew_sarima$rmse), 4)
)

ew_comparison
```

```{r Expanding window forecast plot, fig.width=10, fig.height=5}
ew_df <- tibble(
  date = test_df$date,
  Actual = as.numeric(hh_test),
  `Fixed params`    = as.numeric(fc_store[[best_arima_label]]),
  `Expanding window` = as.numeric(ew_arima$fc)
) |>
  pivot_longer(-date, names_to = "Method", values_to = "value")

ggplot(ew_df, aes(x = date, y = value, color = Method, linetype = Method)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  labs(
    title = paste("Fixed vs expanding window:", best_arima_label),
    x = "Quarter", y = "Household net savings"
  )
```

**Takeaway:** For household net savings, the expanding window RMSE and fixed-parameter RMSE are expected to be similar because (a) the training set is already large relative to the holdout, and (b) the series is close to a random walk, so parameter estimates barely change with 1 additional observation. The near-random-walk behaviour means the primary "forecast" is simply the previous value plus noise, and re-estimating parameters adds marginal value. This further supports the conclusion that the series is difficult to forecast.

## Savings Rate Forecasting Experiment

Instead of forecasting `hh_save` in absolute CAD, we test whether forecasting the **household saving rate** (`hh_rate = hh_save / gdi`) yields better results relative to benchmarks. The saving rate is bounded and may exhibit more stable dynamics.

```{r Savings rate train-test split}
hh_rate_train <- ts(train_df$hh_rate, start = c(1977, 1), frequency = 4)
hh_rate_test  <- ts(test_df$hh_rate,  start = c(2015, 1), frequency = 4)

cat("Saving rate range (train):", round(range(hh_rate_train), 4), "\n")
cat("Saving rate range (test):",  round(range(hh_rate_test), 4), "\n")
```

```{r Savings rate stationarity, fig.width=8, fig.height=6}
dhh_rate_train <- diff(hh_rate_train)

par(mfrow = c(2, 2))
acf(hh_rate_train, lag.max = 20,  main = "ACF: hh_rate")
pacf(hh_rate_train, lag.max = 20, main = "PACF: hh_rate")
acf(dhh_rate_train, lag.max = 20,  main = "ACF: diff(hh_rate)")
pacf(dhh_rate_train, lag.max = 20, main = "PACF: diff(hh_rate)")
par(mfrow = c(1, 1))

cat("ADF test on hh_rate:\n")
print(adf.test(na.omit(as.numeric(hh_rate_train))))
cat("\nADF test on diff(hh_rate):\n")
print(adf.test(na.omit(as.numeric(dhh_rate_train))))
```

```{r Savings rate benchmarks}
rate_rmse_tbl <- tibble(Model = character(), RMSE = double())

# Average
res <- iid_fc(hh_rate_train, hh_rate_test)
rate_rmse_tbl <- add_row(rate_rmse_tbl, Model = "Average", RMSE = res$rmse)

# Persistence
res <- persist_fc(hh_rate_train, hh_rate_test)
rate_rmse_tbl <- add_row(rate_rmse_tbl, Model = "Persistence", RMSE = res$rmse)

# Average by Period
res <- iidbyperiod_fc(hh_rate_train, hh_rate_test, periodicity = 4)
rate_rmse_tbl <- add_row(rate_rmse_tbl, Model = "Average by Period", RMSE = res$rmse)

# SES
hw_ses_rate <- HoltWinters(hh_rate_train, beta = FALSE, gamma = FALSE)
res <- esm_fc(hh_rate_train, hh_rate_test,
              alpha = hw_ses_rate$alpha,
              level = as.numeric(hw_ses_rate$coefficients[1]))
rate_rmse_tbl <- add_row(rate_rmse_tbl, Model = "SES", RMSE = res$rmse)

# Holt Linear
hw_holt_rate <- HoltWinters(hh_rate_train, gamma = FALSE)
res <- lholt_fc(hh_rate_train, hh_rate_test,
                alpha = hw_holt_rate$alpha,
                beta  = hw_holt_rate$beta,
                level = as.numeric(hw_holt_rate$coefficients[1]),
                slope = as.numeric(hw_holt_rate$coefficients[2]))
rate_rmse_tbl <- add_row(rate_rmse_tbl, Model = "Holt Linear", RMSE = res$rmse)

cat("Benchmarks and ESM:\n")
rate_rmse_tbl
```

```{r Savings rate ARIMA}
# auto.arima on saving rate
auto_rate <- auto.arima(hh_rate_train, seasonal = TRUE,
                        stepwise = FALSE, approximation = FALSE)
cat("auto.arima for hh_rate:\n")
summary(auto_rate)

# Test key ARIMA specifications
rate_arima_011 <- arima_one_step_fc(hh_rate_train, hh_rate_test,
                                     order = c(0, 1, 1), include.mean = TRUE)
rate_arima_111 <- arima_one_step_fc(hh_rate_train, hh_rate_test,
                                     order = c(1, 1, 1), include.mean = TRUE)
rate_arima_110 <- arima_one_step_fc(hh_rate_train, hh_rate_test,
                                     order = c(1, 1, 0), include.mean = TRUE)
rate_arima_210 <- arima_one_step_fc(hh_rate_train, hh_rate_test,
                                     order = c(2, 1, 0), include.mean = TRUE)

# SARIMA specifications
rate_sarima_011_002 <- sarima_one_step_fc(hh_rate_train, hh_rate_test,
                                           order = c(0, 1, 1),
                                           seasonal = c(0, 0, 2),
                                           include.mean = TRUE)
rate_sarima_111_100 <- sarima_one_step_fc(hh_rate_train, hh_rate_test,
                                           order = c(1, 1, 1),
                                           seasonal = c(1, 0, 0),
                                           include.mean = TRUE)

rate_arima_tbl <- tibble(
  Model = c("ARIMA(0,1,1)", "ARIMA(1,1,1)", "ARIMA(1,1,0)", "ARIMA(2,1,0)",
            "SARIMA(0,1,1)(0,0,2)[4]", "SARIMA(1,1,1)(1,0,0)[4]"),
  AIC = round(c(rate_arima_011$fit$aic, rate_arima_111$fit$aic,
                rate_arima_110$fit$aic, rate_arima_210$fit$aic,
                rate_sarima_011_002$fit$aic, rate_sarima_111_100$fit$aic), 2),
  RMSE = round(c(rate_arima_011$rmse, rate_arima_111$rmse,
                  rate_arima_110$rmse, rate_arima_210$rmse,
                  rate_sarima_011_002$rmse, rate_sarima_111_100$rmse), 4)
)

rate_arima_tbl
```

```{r Savings rate combined RMSE}
# Add best ARIMA/SARIMA to rate table
best_rate_idx <- which.min(rate_arima_tbl$RMSE)
rate_rmse_tbl <- add_row(rate_rmse_tbl,
                          Model = rate_arima_tbl$Model[best_rate_idx],
                          RMSE = rate_arima_tbl$RMSE[best_rate_idx])

rate_rmse_tbl <- rate_rmse_tbl |>
  mutate(RMSE = round(RMSE, 6)) |>
  arrange(RMSE)

cat("=== Savings Rate RMSE Ranking ===\n")
rate_rmse_tbl

best_rate_model <- rate_rmse_tbl$Model[1]
cat("\nBest model for hh_rate:", best_rate_model, "\n")
cat("Can best model beat Persistence?",
    ifelse(rate_rmse_tbl$RMSE[rate_rmse_tbl$Model == best_rate_model] <
           rate_rmse_tbl$RMSE[rate_rmse_tbl$Model == "Persistence"],
           "YES", "NO"), "\n")
```

```{r Savings rate residual diagnostics, fig.width=10, fig.height=6}
# Diagnostics for the best rate ARIMA model
rate_fits <- list(
  "ARIMA(0,1,1)" = rate_arima_011$fit,
  "ARIMA(1,1,1)" = rate_arima_111$fit,
  "ARIMA(1,1,0)" = rate_arima_110$fit,
  "ARIMA(2,1,0)" = rate_arima_210$fit,
  "SARIMA(0,1,1)(0,0,2)[4]" = rate_sarima_011_002$fit,
  "SARIMA(1,1,1)(1,0,0)[4]" = rate_sarima_111_100$fit
)
best_rate_arima_label <- rate_arima_tbl$Model[best_rate_idx]
cat("Residual diagnostics for", best_rate_arima_label, "on hh_rate:\n")
forecast::checkresiduals(rate_fits[[best_rate_arima_label]])
```

### Savings Rate Findings

The saving rate transformation bounds the response to a more interpretable scale and removes the GDI-driven trend. However, the RMSE ranking above reveals whether this re-framing provides any improvement over naïve benchmarks. If the Persistence model still dominates, the saving rate — like the level series — is also effectively a random walk, confirming that the household savings process is dominated by unpredictable shocks regardless of the scale used.
