---
title: "02_Household_Forecast"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(knitr)
library(tidyverse)
library(forecast)
library(lubridate)
library(ggplot2)
library(tseries)
```

```{r source helpers}
source("../utils/one_step_model.R")
```

## Load Data

We use the train/test splits and differenced training data saved from the EDA notebook directly.

```{r load data}
train_df   <- readRDS("../output/train_df.rds")
test_df    <- readRDS("../output/test_df.rds")
macro_diff <- readRDS("../output/macro_df_diff.rds")

hh_train <- ts(train_df$hh_save, start = c(1977, 1), frequency = 4)
hh_test  <- ts(test_df$hh_save,  start = c(2015, 1), frequency = 4)
```

## ARIMA Order Identification

```{r ACF PACF of hh_save, fig.width=8, fig.height=6}

dhh_train <- diff(hh_train)

par(mfrow = c(2, 2))
acf(hh_train, lag.max = 20, main = "ACF: hh_save")
pacf(hh_train, lag.max = 20, main = "PACF: hh_save")
acf(dhh_train, lag.max = 20, main = "ACF: dhh")
pacf(dhh_train, lag.max = 20, main = "PACF: dhh")
par(mfrow = c(1, 1))
```

- The ACF of `hh_save` decays slowly, confirming non-stationarity and the need for differencing.
- After differencing, the ACF/PACF of `dhh` guides the choice of $p$ and $q$ for ARIMA.
- The candidate models we are considering are
  - ARIMA(1, 1, 1), ARIMA(0, 1, 1), ARIMA(2,1,0)
  - SARIMA(1, 1, 1)(1, 0, 0)[4], SARIMA(1, 1, 1)(0, 0, 1)[4], etc.

## One-Step Benchmark Models

All benchmarks forecast `hh_save` at level directly.

#### IID (Average Forecasts)

```{r IID}
rmse_tbl <- tibble(Model = character(), RMSE = double())
fc_store <- list()

# Average
res <- iid_fc(hh_train, hh_test)
rmse_tbl <- add_row(rmse_tbl, Model = "Average", RMSE = res$rmse)
fc_store$Average <- res$fc

print(res$rmse)
```

```{r IID by year}
# Average by Year
res <- iidbyyear_fc(hh_train, hh_test, periodicity = 4)
rmse_tbl <- add_row(rmse_tbl, Model = "Average by Year", RMSE = res$rmse)
fc_store$`Average by Year` <- res$fc
print(res$rmse)
```

```{r IID by period}
# Average by Period
res <- iidbyperiod_fc(hh_train, hh_test, periodicity = 4)
rmse_tbl <- add_row(rmse_tbl, Model = "Average by Period", RMSE = res$rmse)
fc_store$`Average by Period` <- res$fc
print(res$rmse)
```

### Persistence

```{r Persistence}
# Persistence forecast
res <- persist_fc(hh_train, hh_test)
rmse_tbl <- add_row(rmse_tbl, Model = "Persistence", RMSE = res$rmse)
fc_store$Persistence <- res$fc

print(res$rmse)
```

## Holt-Winters Variants

### SES

```{r Simple exponential smoothing}
# Simple Exponential Smoothing (SES)
hw_ses <- HoltWinters(hh_train, beta = FALSE, gamma = FALSE)
res <- esm_fc(hh_train, hh_test,
              alpha = hw_ses$alpha,
              level = as.numeric(hw_ses$coefficients[1]))
rmse_tbl <- add_row(rmse_tbl, Model = "SES", RMSE = res$rmse)
fc_store$SES <- res$fc

print(res$rmse)
```

### Holt Linear

```{r Holt linear}
hw_holt <- HoltWinters(hh_train, gamma = FALSE)
res <- lholt_fc(hh_train, hh_test,
                alpha = hw_holt$alpha,
                beta  = hw_holt$beta,
                level = as.numeric(hw_holt$coefficients[1]),
                slope = as.numeric(hw_holt$coefficients[2]))
rmse_tbl <- add_row(rmse_tbl, Model = "Holt Linear", RMSE = res$rmse)
fc_store$`Holt Linear` <- res$fc

print(res$rmse)
```

### Holt-Winters Seasonal Additive

```{r Holt-Winters additive}
hw_add <- HoltWinters(hh_train, seasonal = "additive")
res <- aseason_fc(hh_train, hh_test,
                  alpha  = hw_add$alpha,
                  beta   = hw_add$beta,
                  gamma  = hw_add$gamma,
                  level  = as.numeric(hw_add$coefficients[1]),
                  slope  = as.numeric(hw_add$coefficients[2]),
                  season = as.numeric(hw_add$coefficients[3:6]))
                  
rmse_tbl <- add_row(rmse_tbl, Model = "HW Additive", RMSE = res$rmse)
fc_store$`HW Additive` <- res$fc

print(res$rmse)
```

## ARIMA One-Step Forecast

```{r ARIMA one-step forecast}
# Safely extract ARIMA order
extract_order <- function(ord, keys, default = 0L) {
  vals <- ord[keys]
  vals[is.na(vals)] <- default
  as.integer(vals)
}

fit_arima <- auto.arima(hh_train,
                        seasonal = TRUE,
                        stepwise = FALSE,
                        approximation = FALSE)
# Print Auto ARIMA summary                   
summary(fit_arima)
```

```{r Alternative ARIMA specifications}
# Try alternative ARIMA specifications to check robustness of order selection
fit_arima_111 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  include.mean = TRUE
)

fit_arima_011 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  include.mean = TRUE
)

fit_arima_210 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(2, 1, 0),
  include.mean = TRUE
)

fit_arima_110 <- arima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 0),
  include.mean = TRUE
)

arima_tbl <- tibble(
  Model = c("ARIMA(1,1,1)", "ARIMA(0,1,1)", "ARIMA(2,1,0)", "ARIMA(1,1,0)"),
  RMSE = c(fit_arima_111$rmse, fit_arima_011$rmse, 
           fit_arima_210$rmse, fit_arima_110$rmse)
)

arima_tbl
```
```{r Add the best ARIMA}
rmse_tbl <- add_row(rmse_tbl, Model = "ARIMA(1,1,1)", RMSE = min(arima_tbl$RMSE))

fc_store$`ARIMA(1,1,1)` <- fit_arima_111$fc
```

```{r Auto SARIMA}
ord <- forecast::arimaorder(fit_arima)
res <- sarima_one_step_fc(
  train    = hh_train,
  holdout  = hh_test,
  order    = extract_order(ord, c("p", "d", "q")),
  seasonal = extract_order(ord, c("P", "D", "Q")),
  include.mean = TRUE
)
rmse_tbl <- add_row(rmse_tbl, Model = "Auto ARIMA", RMSE = res$rmse)
fc_store$`Auto ARIMA` <- res$fc
print(res$rmse)
```

```{r Try alternative SARIMA specifications}
fit_sarima_111_100 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(1, 0, 0),
  include.mean = TRUE
)

fit_sarima_011_100 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(1, 0, 0),
  include.mean = TRUE
)

fit_sarima_111_101 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(1, 0, 1),
  include.mean = TRUE
)

fit_sarima_011_101 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(1, 0, 1),
  include.mean = TRUE
)

fit_sarima_111_001 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(0, 0, 1),
  include.mean = TRUE
)

fit_sarima_011_001 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(0, 0, 1),
  include.mean = TRUE
)

fit_sarima_111_002 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(1, 1, 1),
  seasonal = c(0, 0, 2),
  include.mean = TRUE
)

fit_sarima_011_002 <- sarima_one_step_fc(
  train = hh_train,
  holdout = hh_test,
  order = c(0, 1, 1),
  seasonal = c(0, 0, 2),
  include.mean = TRUE
)

sarima_tbl <- tibble(
  Model = c(
    "SARIMA(1,1,1)(1,0,0)[4]",
    "SARIMA(0,1,1)(1,0,0)[4]",
    "SARIMA(1,1,1)(1,0,1)[4]",
    "SARIMA(0,1,1)(1,0,1)[4]",
    "SARIMA(1,1,1)(0,0,1)[4]",
    "SARIMA(0,1,1)(0,0,1)[4]",
    "SARIMA(1,1,1)(0,0,2)[4]",
    "SARIMA(0,1,1)(0,0,2)[4]"
  ),
  RMSE = c(
    fit_sarima_111_100$rmse,
    fit_sarima_011_100$rmse,
    fit_sarima_111_101$rmse,
    fit_sarima_011_101$rmse,
    fit_sarima_111_001$rmse,
    fit_sarima_011_001$rmse,
    fit_sarima_111_002$rmse,
    fit_sarima_011_002$rmse
  )
)

sarima_tbl
```

```{r Store the best SARIMA}
rmse_tbl <- add_row(rmse_tbl, Model = "SARIMA(1,1,1)(1,0,0)[4]", 
                    RMSE = min(sarima_tbl$RMSE))
fc_store$`SARIMA(1,1,1)(1,0,0)[4]` <- fit_sarima_111_100$fc
```

## ARIMAX One-Step Forecast

### Exogenous Variable Selection via CCF

We use the differenced training data from the EDA to inspect cross-correlations with `dhh`. A variable is selected for ARIMAX only if it shows significant correlation at **negative lags** (the regressor leads `dhh`), to avoid overfitting with only 20 holdout observations.

```{r CCF variable selection, fig.width=10, fig.height=6}
candidates <- c("dcorp", "ldgdi", "ldcons", "dunemp", "dinterest", "ldcpi", "ldgpdi")

par(
  mfrow = c(2, 4),
  mar = c(5, 5, 3, 1),
  mgp = c(2, 0.8, 0),
  cex.main = 0.95,
  cex.lab = 0.9,
  cex.axis = 0.85
)
for (v in candidates) {
  ccf(na.omit(macro_diff[[v]]), na.omit(macro_diff$dhh),
      lag.max = 16,
      main = paste(v, "vs dhh"))
}
par(mfrow = c(1, 1))
```

```{r CCF values}
ccf_values <- sapply(candidates, function(v) {
  ccf_res <- ccf(na.omit(macro_diff[[v]]), na.omit(macro_diff$dhh), lag.max = 16, plot = FALSE)
  max_ccf <- max(abs(ccf_res$acf[ccf_res$lag < 0]))
  return(max_ccf)
})

# Identify the lag with the maximum absolute CCF for each candidate
ccf_lags <- sapply(candidates, function(v) {
  ccf_res <- ccf(na.omit(macro_diff[[v]]), na.omit(macro_diff$dhh), lag.max = 16, plot = FALSE)
  max_idx <- which.max(abs(ccf_res$acf[ccf_res$lag < 0]))
  return(ccf_res$lag[max_idx])
})

ccf_tbl <- tibble(
  Variable = candidates,
  Max_CCF = ccf_values,
  Lag_at_Max_CCF = ccf_lags
)

ccf_tbl
```

Based on the CCF evidence and the EDA findings, we select regressors with clear leading relationship:

- `ldgpdi`
- `ldcpi`

```{r ARIMAX preprocessing}
# Compute stationary exogenous variables across the full sample
arimax_df <- bind_rows(train_df, test_df) |>
  arrange(date) |>
  mutate(
    ldgpdi  = log(gpdi) - log(lag(gpdi)),
    ldcpi = log1p(cpi) - log1p(lag(cpi)),
    ldcpi_lag2 = lag(ldcpi, 2),
    ldgpdi_lag1 = lag(ldgpdi, 1),
    ldgpdi_lag3 = lag(ldgpdi, 3)
  ) |>
  drop_na(ldgpdi_lag1, ldgpdi_lag3) # Drop initial rows with NA due to differencing and lagging

arimax_vars <- c("ldgpdi_lag1", "ldcpi_lag2", "ldgpdi_lag3")

x_train <- arimax_df |> filter(date <= as.Date("2014-12-31")) |>
  select(all_of(arimax_vars)) |>
  as.matrix()
x_test <- arimax_df |> filter(date > as.Date("2014-12-31")) |>
  select(all_of(arimax_vars)) |>
  as.matrix()
```

### Fitting ARIMAX models

```{r ARIMAX one-step forecast}

# Align hh_save accordingly
hh_train_x <- ts(
  arimax_df |> filter(date <= as.Date("2014-12-31")) |> pull(hh_save),
  start = c(
    year(min(arimax_df$date)),
    quarter(min(arimax_df$date))
  ), frequency = 4
)

# Fit ARIMAX with selected regressors
fit_arimax <- auto.arima(hh_train_x,
                         xreg = x_train,
                         seasonal = FALSE,
                         stepwise = FALSE,
                         approximation = FALSE)
summary(fit_arimax)
```

```{r ARIMAX forecast}
arimax_213 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(2, 1, 3),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_212 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(2, 1, 2),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_112 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 2),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_111 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 1),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_011 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(0, 1, 1),
  xreg_train = x_train,
  xreg_holdout = x_test,
  include.mean = TRUE
)

arimax_tbl <- tibble(
  Model = c(
    "ARIMAX(2,1,3)",
    "ARIMAX(2,1,2)",
    "ARIMAX(1,1,2)",
    "ARIMAX(1,1,1)",
    "ARIMAX(0,1,1)"
  ),
  RMSE = c(
    arimax_213$rmse,
    arimax_212$rmse,
    arimax_112$rmse,
    arimax_111$rmse,
    arimax_011$rmse
  )
)

arimax_tbl
```

## Theoretical Model (Precautionary Saving)

In Macroeconomics, the precautionary saving motive is modeled by the permanent income model. Where, assuming a representative household is constant relative risk averse, we have:
$$
C_t^{-\gamma} = \beta\mathbb{E}_t[(1+r_{t+1})C_{t+1}] 
$$
where $C_t$ is consumption, $\beta$ is the discount factor, $\gamma$ is the coefficient of relative risk aversion, and $r_{t+1}$ is the real interest rate. This implies that consumption growth depends on expected returns and the intertemporal elasticity of substitution.

Taking the log-transformation and Taylor expansion will eventually give us
$$
c_t = c_{t+1} - \frac{1}{\gamma} \ln \beta - \frac{1}{\gamma} r_{t+1}  - \frac{\gamma}{2} \text{Var}_t(\Delta c_{t+1}) - \varepsilon_{t+1}
$$
Where with the identity $s_t = y_t - c_t$ , the first difference of saving is captured by
$$
\Delta s_t = \Delta y_t - \Delta\mathbb{E}_t[c_{t+1}] + \frac{1}{\gamma}\Delta r_{t+1} + \frac{\gamma}{2} \Delta\text{Var}_t(\Delta c_{t+1}) + \varepsilon_{t+1}
$$
This suggests we can fit the following empirical model
$$
dhh_t = \alpha + \sum_{i=0}^p \phi_i dhh_{t-i} + \beta_1 dinterest_{t-1} + \beta_2 dunemp_{t-1} + \beta_3 ldgdi_{t-1} + \beta_4 ldgpdi_{t-1} + \varepsilon_t
$$
Where we use the unemployment and Gross Private Domestic Investment (GPDI) as proxies for uncertainty, and the real interest rate as the return variable. We lag all regressors by 1 quarter to avoid look-ahead bias in forecasting.

```{r Rainy day forecast}
pim_df <- bind_rows(train_df, test_df) |>
  arrange(date) |>
  mutate(
    dinterest = interest - lag(interest),
    dunemp = unemp - lag(unemp),
    ldgdi = log(gdi) - log(lag(gdi)),
    ldgpdi = log(gpdi) - log(lag(gpdi)),
    dinterest_lag1 = lag(dinterest, 1),
    dunemp_lag1 = lag(dunemp, 1),
    ldgdi_lag1 = lag(ldgdi, 1),
    ldgpdi_lag1 = lag(ldgpdi, 1)
  ) |>
  drop_na(ldgdi_lag1)

pim_train <- pim_df |> filter(date <= as.Date("2014-12-31"))|>
  select(dinterest_lag1, dunemp_lag1, ldgdi_lag1, ldgpdi_lag1)|>
  as.matrix()
pim_test <- pim_df |> filter(date > as.Date("2014-12-31"))|>
  select(dinterest_lag1, dunemp_lag1, ldgdi_lag1, ldgpdi_lag1)|>
  as.matrix()

hh_train_x <- ts(
  pim_df |> filter(date <= as.Date("2014-12-31"))|>pull(hh_save),
  start = c(
    year(min(pim_df$date)),
    quarter(min(pim_df$date))
  ), 
  frequency = 4
)
```

```{r Fit the rainy day model}
pim_fit <- auto.arima(hh_train_x, 
                      xreg = pim_train,
                      seasonal = FALSE,
                      stepwise = FALSE,
                      approximation = FALSE)
pim_fit
```
```{r Test alternatives}
pim_111 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 1),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)

pim_011 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(0, 1, 1),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)


pim_110 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(1, 1, 0),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)

pim_012 <- arimax_one_step_fc(
  train = hh_train_x,
  holdout = hh_test,
  order = c(0, 1, 2),
  xreg_train = pim_train,
  xreg_holdout = pim_test,
  include.mean = TRUE
)

pim_tbl <- tibble(
  Model = c(
    "PIM ARIMAX(1,1,1)",
    "PIM ARIMAX(0,1,1)",
    "PIM ARIMAX(1,1,0)",
    "PIM ARIMAX(0,1,2)"
  ),
  RMSE = c(
    pim_111$rmse,
    pim_011$rmse,
    pim_110$rmse,
    pim_012$rmse
  )
)

pim_tbl
```

```{r Store the best PIM ARIMAX}
rmse_tbl <- add_row(rmse_tbl, Model = "PIM ARIMAX(0,1,2)",
                    RMSE = min(pim_tbl$RMSE))
fc_store$`PIM ARIMAX(0,1,2)` <- pim_012$fc
```

## RMSE Ranking

```{r RMSE ranking}
rmse_tbl <- rmse_tbl |>
  mutate(RMSE = round(RMSE, 4)) |>
  arrange(RMSE)

rmse_tbl

best_model <- rmse_tbl$Model[1]
cat("Best model by one-step holdout RMSE:", best_model, "\n")
```

## Holdout Visualization

```{r holdout forecast plot, fig.width=10, fig.height=6}
fc_df <- tibble(date = test_df$date, Actual = as.numeric(hh_test))


for (m in names(fc_store)) {
  fc_df[[m]] <- as.numeric(fc_store[[m]])
}

fc_long <- fc_df |>
  pivot_longer(cols = -date, names_to = "Model", values_to = "value")|>
  filter(
    Model %in% c("Actual", "Average", "Persistence", 
                 "HW Additive", 
                 "ARIMA(1,1,1)", 
                 "SARIMA(1,1,1)(1,0,0)[4]",
                 "PIM ARIMAX(0,1,2)")
  ) # Keep only benchmarks and the best HW and ARIMA/SARIMA models for clarity

ggplot(fc_long, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "One-step-ahead holdout forecasts of household net savings",
    x = "Quarter",
    y = "Household net savings",
    color = "Model",
    linetype = "Model"
  )
```

## Interpretation

- All benchmarks and ARIMA/ARIMAX models forecast `hh_save` directly; differencing is handled internally by `auto.arima` via the $d$ parameter.
- The ARIMAX uses only CCF-justified regressors (`r paste(arimax_vars, collapse = ", ")`) to avoid overfitting the small holdout window.
- The RMSE ranking table determines the best model for household net savings forecasting.
