---
title: "03_Expanding_Window_Forecast"
output:
  pdf_document: default
  html_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(knitr)
library(tidyverse)
library(forecast)
library(lubridate)
library(ggplot2)
library(tseries)
```

```{r source helpers}
source("../utils/one_step_model.R")
source("../utils/expanding_window.R")
```

## Load Data

We use the full EDA dataset (pre-differenced) saved from notebook 01.

```{r load data}
eda_df <- readRDS("../output/eda_df_nodiff.rds")

full_df <- eda_df |> filter(date <= as.Date("2019-12-31"))

hh_full <- ts(full_df$hh_save, start = c(1977, 1), frequency = 4)
```

## Expanding Window Design

We use the first **20 years** (80 quarters, 1977 Q1 – 1996 Q4) as the initial training window.
The expanding window then re-estimates the model at each subsequent quarter:

- Step 1: Train on 1977 Q1 – 1996 Q4, forecast 1997 Q1.
- Step 2: Train on 1977 Q1 – 1997 Q1, forecast 1997 Q2.
- Continue until the end of the series (2019 Q4).

This yields **92 one-step-ahead forecasts** (1997 Q1 – 2019 Q4), each produced by a model fitted to an ever-growing training window.

```{r train-test split for expanding window}
initial_end <- as.Date("1996-12-31")

ew_train_df <- full_df |> filter(date <= initial_end)
ew_test_df  <- full_df |> filter(date > initial_end)

hh_train <- ts(ew_train_df$hh_save, start = c(1977, 1), frequency = 4)
hh_test  <- ts(ew_test_df$hh_save,  start = c(1997, 1), frequency = 4)

cat("Initial training window:", nrow(ew_train_df))
cat("Holdout window:", nrow(ew_test_df))
```

```{r save expanding window splits}
saveRDS(ew_train_df, file = "../output/ew_train_df.rds")
saveRDS(ew_test_df,  file = "../output/ew_test_df.rds")
```

## ARIMA Order Identification

```{r ACF PACF, fig.width=8, fig.height=6}
dhh_train <- diff(hh_train)

par(mfrow = c(2, 2))
acf(hh_train, lag.max = 20, main = "ACF: hh_save (train)")
pacf(hh_train, lag.max = 20, main = "PACF: hh_save (train)")
acf(dhh_train, lag.max = 20, main = "ACF: diff(hh_save)")
pacf(dhh_train, lag.max = 20, main = "PACF: diff(hh_save)")
par(mfrow = c(1, 1))
```

```{r auto.arima on initial window}
fit_auto <- auto.arima(hh_train,
                       seasonal = TRUE,
                       stepwise = FALSE,
                       approximation = FALSE)
summary(fit_auto)
```

## Training Set Fit

Before evaluating holdout performance, we inspect how each model fits the initial 80-quarter training set. This diagnostic reveals whether a model can capture the in-sample dynamics of household net savings.

### Benchmark Training Fits

```{r benchmark training fits, fig.width=10, fig.height=8}
n_train <- length(hh_train)
train_dates <- ew_train_df$date

# Persistence: fitted[t] = y[t-1]
persist_train_fitted <- c(NA, as.numeric(hh_train[-n_train]))

# Average: fitted[t] = overall mean of training set
avg_train_fitted <- rep(mean(hh_train), n_train)

# SES
hw_ses <- HoltWinters(hh_train, beta = FALSE, gamma = FALSE)
ses_fitted_raw <- as.numeric(hw_ses$fitted[, "xhat"])
ses_train_fitted <- c(rep(NA, n_train - length(ses_fitted_raw)), ses_fitted_raw)

# Holt Linear
hw_holt <- HoltWinters(hh_train, gamma = FALSE)
holt_fitted_raw <- as.numeric(hw_holt$fitted[, "xhat"])
holt_train_fitted <- c(rep(NA, n_train - length(holt_fitted_raw)), holt_fitted_raw)

# HW Additive
hw_add <- HoltWinters(hh_train, seasonal = "additive")
hwa_fitted_raw <- as.numeric(hw_add$fitted[, "xhat"])
hwa_train_fitted <- c(rep(NA, n_train - length(hwa_fitted_raw)), hwa_fitted_raw)

train_fit_df <- tibble(
  date = train_dates,
  Actual = as.numeric(hh_train),
  Persistence = persist_train_fitted,
  Average = avg_train_fitted,
  SES = ses_train_fitted,
  `Holt Linear` = holt_train_fitted,
  `HW Additive` = hwa_train_fitted
) |>
  pivot_longer(-date, names_to = "Model", values_to = "value")

ggplot(train_fit_df, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Training set fit: Benchmark models (1977 Q1 – 1996 Q4)",
    x = "Quarter", y = "Household net savings"
  )
```

### ARIMA / SARIMA Training Fits

```{r ARIMA SARIMA training fits, fig.width=10, fig.height=8}
# Fit each ARIMA / SARIMA on the initial training window
arima_specs <- list(
  "ARIMA(1,1,1)" = c(1, 1, 1),
  "ARIMA(0,1,1)" = c(0, 1, 1),
  "ARIMA(2,1,0)" = c(2, 1, 0),
  "ARIMA(1,1,0)" = c(1, 1, 0)
)

sarima_specs <- list(
  "SARIMA(1,1,1)(1,0,0)[4]" = list(order = c(1,1,1), seasonal = c(1,0,0)),
  "SARIMA(0,1,1)(1,0,0)[4]" = list(order = c(0,1,1), seasonal = c(1,0,0)),
  "SARIMA(1,1,1)(0,0,1)[4]" = list(order = c(1,1,1), seasonal = c(0,0,1)),
  "SARIMA(0,1,1)(0,0,1)[4]" = list(order = c(0,1,1), seasonal = c(0,0,1))
)

# Store training fits for plotting
arima_train_fits <- list()
for (nm in names(arima_specs)) {
  fit_obj <- forecast::Arima(hh_train, order = arima_specs[[nm]], include.mean = TRUE)
  arima_train_fits[[nm]] <- as.numeric(fitted(fit_obj))
}

sarima_train_fits <- list()
for (nm in names(sarima_specs)) {
  sp <- sarima_specs[[nm]]
  fit_obj <- forecast::Arima(hh_train, order = sp$order, seasonal = sp$seasonal,
                             include.mean = TRUE)
  sarima_train_fits[[nm]] <- as.numeric(fitted(fit_obj))
}

arima_train_df <- tibble(date = train_dates, Actual = as.numeric(hh_train))
for (nm in names(arima_train_fits)) {
  arima_train_df[[nm]] <- arima_train_fits[[nm]]
}
for (nm in names(sarima_train_fits)) {
  arima_train_df[[nm]] <- sarima_train_fits[[nm]]
}

arima_train_long <- arima_train_df |>
  pivot_longer(-date, names_to = "Model", values_to = "value")

ggplot(arima_train_long, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Training set fit: ARIMA / SARIMA models (1977 Q1 – 1996 Q4)",
    x = "Quarter", y = "Household net savings"
  )
```

```{r training RMSE comparison}
train_rmse <- function(actual, fitted) {
  valid <- !is.na(fitted)
  sqrt(mean((actual[valid] - fitted[valid])^2))
}

train_rmse_tbl <- tibble(
  Model = c("Persistence", "Average", "SES", "Holt Linear", "HW Additive",
            names(arima_train_fits), names(sarima_train_fits)),
  Train_RMSE = round(c(
    train_rmse(as.numeric(hh_train), persist_train_fitted),
    train_rmse(as.numeric(hh_train), avg_train_fitted),
    train_rmse(as.numeric(hh_train), ses_train_fitted),
    train_rmse(as.numeric(hh_train), holt_train_fitted),
    train_rmse(as.numeric(hh_train), hwa_train_fitted),
    sapply(arima_train_fits, \(f) train_rmse(as.numeric(hh_train), f)),
    sapply(sarima_train_fits, \(f) train_rmse(as.numeric(hh_train), f))
  ), 4)
) |>
  arrange(Train_RMSE)

# Print training RMSE table
train_rmse_tbl
```

## Benchmark Holdout Forecasts

We evaluate all benchmark rules on the 92-quarter holdout for reference.

```{r benchmarks}
ew_rmse_tbl <- tibble(Model = character(), Method = character(), RMSE = double())
ew_fc_store <- list()

# Persistence
res <- persist_fc(hh_train, hh_test)
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Persistence", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$Persistence <- res$fc

# Average
res <- iid_fc(hh_train, hh_test)
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Average", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$Average <- res$fc

# Average by Period
res <- iidbyperiod_fc(hh_train, hh_test, periodicity = 4)
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Average by Period", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$`Average by Period` <- res$fc

# SES
res <- esm_fc(hh_train, hh_test,
              alpha = hw_ses$alpha,
              level = as.numeric(hw_ses$coefficients[1]))
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "SES", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$SES <- res$fc

# Holt Linear
res <- lholt_fc(hh_train, hh_test,
                alpha = hw_holt$alpha,
                beta  = hw_holt$beta,
                level = as.numeric(hw_holt$coefficients[1]),
                slope = as.numeric(hw_holt$coefficients[2]))
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Holt Linear", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$`Holt Linear` <- res$fc

# HW Additive
res <- aseason_fc(hh_train, hh_test,
                  alpha  = hw_add$alpha,
                  beta   = hw_add$beta,
                  gamma  = hw_add$gamma,
                  level  = as.numeric(hw_add$coefficients[1]),
                  slope  = as.numeric(hw_add$coefficients[2]),
                  season = as.numeric(hw_add$coefficients[3:6]))
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "HW Additive", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$`HW Additive` <- res$fc

cat("Benchmark RMSEs:", ew_rmse_tbl)
```

## Fixed-Parameter One-Step Forecasts (ARIMA / SARIMA)

For reference, we run fixed-parameter Kalman filtering (same approach as notebook 02) on this longer holdout.

```{r ARIMA fixed parameter}
arima_fixed_results <- list()

for (nm in names(arima_specs)) {
  res <- arima_one_step_fc(hh_train, hh_test,
                           order = arima_specs[[nm]],
                           include.mean = TRUE)
  arima_fixed_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Fixed", RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (fixed)")]] <- res$fc
}

arima_fixed_tbl <- tibble(
  Model = names(arima_specs),
  AIC = sapply(arima_fixed_results, \(r) round(r$fit$aic, 2)),
  RMSE_Fixed = sapply(arima_fixed_results, \(r) round(r$rmse, 4))
)
arima_fixed_tbl
```

```{r SARIMA fixed parameter}
sarima_fixed_results <- list()
for (nm in names(sarima_specs)) {
  sp <- sarima_specs[[nm]]
  res <- sarima_one_step_fc(hh_train, hh_test,
                            order = sp$order,
                            seasonal = sp$seasonal,
                            include.mean = TRUE)
  sarima_fixed_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Fixed", RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (fixed)")]] <- res$fc
}

sarima_fixed_tbl <- tibble(
  Model = names(sarima_specs),
  AIC = sapply(sarima_fixed_results, \(r) round(r$fit$aic, 2)),
  RMSE_Fixed = sapply(sarima_fixed_results, \(r) round(r$rmse, 4))
)
sarima_fixed_tbl
```

## Expanding Window Forecasts

Now we re-estimate each model at every holdout step.

```{r ARIMA expanding window}
arima_ew_results <- list()
for (nm in names(arima_specs)) {
  res <- arima_expanding_window_fc(hh_train, hh_test,
                                   order = arima_specs[[nm]],
                                   include.mean = TRUE)
  arima_ew_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Expanding", 
                         RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (expanding)")]] <- res$fc
}
```

```{r SARIMA expanding window}
sarima_ew_results <- list()
for (nm in names(sarima_specs)) {
  sp <- sarima_specs[[nm]]
  res <- sarima_expanding_window_fc(hh_train, hh_test,
                                    order = sp$order,
                                    seasonal = sp$seasonal,
                                    include.mean = TRUE)
  sarima_ew_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Expanding", 
                         RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (expanding)")]] <- res$fc
}
```

## Comparison Tables

### Fixed vs Expanding Window

```{r comparison table}
# ARIMA: fixed vs expanding
arima_compare <- tibble(
  Model = names(arima_specs),
  RMSE_Fixed = sapply(arima_fixed_results, \(r) round(r$rmse, 4)),
  RMSE_Expanding = sapply(arima_ew_results, \(r) round(r$rmse, 4)),
  Improvement_pct = round((RMSE_Fixed - RMSE_Expanding) / RMSE_Fixed * 100, 2)
)
cat("=== ARIMA: Fixed vs Expanding ===\n")
arima_compare

# SARIMA: fixed vs expanding
sarima_compare <- tibble(
  Model = names(sarima_specs),
  RMSE_Fixed = sapply(sarima_fixed_results, \(r) round(r$rmse, 4)),
  RMSE_Expanding = sapply(sarima_ew_results, \(r) round(r$rmse, 4)),
  Improvement_pct = round((RMSE_Fixed - RMSE_Expanding) / RMSE_Fixed * 100, 2)
)
cat("\n=== SARIMA: Fixed vs Expanding ===\n")
sarima_compare
```

### Full RMSE Ranking (Benchmarks + ARIMA/SARIMA)

```{r overall RMSE ranking}
ew_rmse_tbl <- ew_rmse_tbl |>
  mutate(RMSE = round(RMSE, 4)) |>
  arrange(RMSE)

ew_rmse_tbl

best_ew_model <- ew_rmse_tbl |> slice(1)

best_ew_model
```

```{r persistence comparison}
# How do expanding window models compare to persistence?
persist_rmse <- ew_rmse_tbl |> filter(Model == "Persistence") |> pull(RMSE)


cat("Models that beat Persistence:\n")
ew_rmse_tbl |> filter(RMSE < persist_rmse)
```

```{r save expanding window results}
saveRDS(ew_rmse_tbl, file = "../output/ew_rmse_tbl.rds")
```

## Holdout Visualization

### Benchmarks vs Best Expanding Window Models

```{r all models holdout, fig.width=10, fig.height=6}
best_arima_nm  <- arima_compare$Model[which.min(arima_compare$RMSE_Expanding)]
best_sarima_nm <- sarima_compare$Model[which.min(sarima_compare$RMSE_Expanding)]

fc_all_df <- tibble(
  date = ew_test_df$date,
  Actual = as.numeric(hh_test),
  Persistence = as.numeric(ew_fc_store$Persistence),
  Average = as.numeric(ew_fc_store$Average),
  SES = as.numeric(ew_fc_store$SES),
  `Holt Linear` = as.numeric(ew_fc_store$`Holt Linear`),
  `HW Additive` = as.numeric(ew_fc_store$`HW Additive`)
)
fc_all_df[[paste0(best_arima_nm, " (exp)")]] <-
  as.numeric(ew_fc_store[[paste0(best_arima_nm, " (expanding)")]])
fc_all_df[[paste0(best_sarima_nm, " (exp)")]] <-
  as.numeric(ew_fc_store[[paste0(best_sarima_nm, " (expanding)")]])

fc_all_long <- fc_all_df |>
  pivot_longer(-date, names_to = "Model", values_to = "value")

ggplot(fc_all_long, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Holdout forecasts: Benchmarks vs best expanding window models",
    x = "Quarter", y = "Household net savings"
  )
```

### Fixed vs Expanding: Best ARIMA

```{r best ARIMA fixed vs expanding, fig.width=10, fig.height=5}
fc_df <- tibble(
  date = ew_test_df$date,
  Actual = as.numeric(hh_test),
  `Fixed params`     = as.numeric(ew_fc_store[[paste0(best_arima_nm, " (fixed)")]]),
  `Expanding window`  = as.numeric(ew_fc_store[[paste0(best_arima_nm, " (expanding)")]]),
  Persistence        = as.numeric(ew_fc_store$Persistence)
) |>
  pivot_longer(-date, names_to = "Method", values_to = "value")

ggplot(fc_df, aes(x = date, y = value, color = Method, linetype = Method)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = paste("Fixed vs Expanding Window:", best_arima_nm),
    x = "Quarter", y = "Household net savings"
  )
```

### Fixed vs Expanding: Best SARIMA

```{r best SARIMA fixed vs expanding, fig.width=10, fig.height=5}
fc_df2 <- tibble(
  date = ew_test_df$date,
  Actual = as.numeric(hh_test),
  `Fixed params`     = as.numeric(ew_fc_store[[paste0(best_sarima_nm, " (fixed)")]]),
  `Expanding window`  = as.numeric(ew_fc_store[[paste0(best_sarima_nm, " (expanding)")]]),
  Persistence        = as.numeric(ew_fc_store$Persistence)
) |>
  pivot_longer(-date, names_to = "Method", values_to = "value")

ggplot(fc_df2, aes(x = date, y = value, color = Method, linetype = Method)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = paste("Fixed vs Expanding Window:", best_sarima_nm),
    x = "Quarter", y = "Household net savings"
  )
```

## Interpretation

- The expanding window re-estimates model parameters at each step, allowing them to adapt to new data. Over a 92-quarter holdout, any structural shifts in the savings process should be captured more faithfully than with parameters fixed at 1996 Q4 values.
- Comparing the expanding window ARIMA/SARIMA models directly against benchmark rules (Persistence, Average, SES, Holt, HW Additive) reveals whether the added complexity of parametric models is justified.
- If the expanding window RMSE is similar to the fixed-parameter RMSE, this confirms that the series is close to a random walk, and parameter re-estimation adds little value.
- The persistence benchmark remains the key reference: any model must beat persistence to justify its complexity.
- The training set fit diagnostic helps identify whether a model captures the historical dynamics or merely overfits. A model with low training RMSE but high holdout RMSE is likely overfitting.
