---
title: "03_Rolling_Window_Forecast"
output:
  pdf_document: default
  html_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(knitr)
library(tidyverse)
library(forecast)
library(lubridate)
library(ggplot2)
library(tseries)
```

```{r source helpers}
source("../utils/one_step_model.R")
source("../utils/expanding_window.R")
```

## Load Data

We use the full EDA dataset (pre-differenced) saved from notebook 01.

```{r load data}
eda_df <- readRDS("../output/eda_df_nodiff.rds")

full_df <- eda_df |> filter(date <= as.Date("2019-12-31"))

hh_full <- ts(full_df$hh_save, start = c(1977, 1), frequency = 4)
```

## Rolling Window Design

Economic time series that span multiple regimes (e.g., oil shocks, the Great
Moderation, the 2008 financial crisis) are prone to structural breaks. An
expanding window assigns equal weight to all past observations, so data from an
old regime can dilute parameter estimates for the current one. A **fixed rolling
window** of size $R$ discards observations older than $R$ quarters, keeping
estimation sample size constant and allowing the model to adapt to regime
changes (Giacomini & White, 2006).

We set $R = 80$ quarters (20 years), matching the initial training length.

- Step 1: Train on 1977 Q1 -- 1996 Q4 (80 obs), forecast 1997 Q1.
- Step 2: Train on 1977 Q2 -- 1997 Q1 (80 obs), forecast 1997 Q2.
- Continue until the end of the series (2019 Q4).

This yields **92 one-step-ahead forecasts** (1997 Q1 -- 2019 Q4), each produced
by a model fitted to the most recent 80 quarters.

```{r rolling window parameters}
R_window <- 80  # 20 years * 4 quarters
```

```{r train-test split for rolling window}
initial_end <- as.Date("1996-12-31")

ew_train_df <- full_df |> filter(date <= initial_end)
ew_test_df  <- full_df |> filter(date > initial_end)

hh_train <- ts(ew_train_df$hh_save, start = c(1977, 1), frequency = 4)
hh_test  <- ts(ew_test_df$hh_save,  start = c(1997, 1), frequency = 4)

cat("Initial training window:", nrow(ew_train_df))
cat("Holdout window:", nrow(ew_test_df))
```

```{r save rolling window splits}
saveRDS(ew_train_df, file = "../output/ew_train_df.rds")
saveRDS(ew_test_df,  file = "../output/ew_test_df.rds")
```

## ARIMA Order Identification

```{r ACF PACF, fig.width=8, fig.height=6}
dhh_train <- diff(hh_train)

par(mfrow = c(2, 2))
acf(hh_train, lag.max = 20, main = "ACF: hh_save (train)")
pacf(hh_train, lag.max = 20, main = "PACF: hh_save (train)")
acf(dhh_train, lag.max = 20, main = "ACF: diff(hh_save)")
pacf(dhh_train, lag.max = 20, main = "PACF: diff(hh_save)")
par(mfrow = c(1, 1))
```

```{r auto.arima on initial window}
fit_auto <- auto.arima(hh_train,
                       seasonal = TRUE,
                       stepwise = FALSE,
                       approximation = FALSE)
summary(fit_auto)
```

## Training Set Fit

Before evaluating holdout performance, we inspect how each model fits the initial 80-quarter training set. This diagnostic reveals whether a model can capture the in-sample dynamics of household net savings.

### Benchmark Training Fits

```{r benchmark training fits, fig.width=10, fig.height=8}
n_train <- length(hh_train)
train_dates <- ew_train_df$date

# Persistence: fitted[t] = y[t-1]
persist_train_fitted <- c(NA, as.numeric(hh_train[-n_train]))

# Average: fitted[t] = overall mean of training set
avg_train_fitted <- rep(mean(hh_train), n_train)

# SES
hw_ses <- HoltWinters(hh_train, beta = FALSE, gamma = FALSE)
ses_fitted_raw <- as.numeric(hw_ses$fitted[, "xhat"])
ses_train_fitted <- c(rep(NA, n_train - length(ses_fitted_raw)), ses_fitted_raw)

# Holt Linear
hw_holt <- HoltWinters(hh_train, gamma = FALSE)
holt_fitted_raw <- as.numeric(hw_holt$fitted[, "xhat"])
holt_train_fitted <- c(rep(NA, n_train - length(holt_fitted_raw)), holt_fitted_raw)

# HW Additive
hw_add <- HoltWinters(hh_train, seasonal = "additive")
hwa_fitted_raw <- as.numeric(hw_add$fitted[, "xhat"])
hwa_train_fitted <- c(rep(NA, n_train - length(hwa_fitted_raw)), hwa_fitted_raw)

train_fit_df <- tibble(
  date = train_dates,
  Actual = as.numeric(hh_train),
  Persistence = persist_train_fitted,
  Average = avg_train_fitted,
  SES = ses_train_fitted,
  `Holt Linear` = holt_train_fitted,
  `HW Additive` = hwa_train_fitted
) |>
  pivot_longer(-date, names_to = "Model", values_to = "value")

ggplot(train_fit_df, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Training set fit: Benchmark models (1977 Q1 -- 1996 Q4)",
    x = "Quarter", y = "Household net savings"
  )
```

### ARIMA / SARIMA Training Fits

```{r ARIMA SARIMA training fits, fig.width=10, fig.height=8}
# Fit each ARIMA / SARIMA on the initial training window
arima_specs <- list(
  "ARIMA(1,1,1)" = c(1, 1, 1),
  "ARIMA(0,1,1)" = c(0, 1, 1),
  "ARIMA(2,1,0)" = c(2, 1, 0),
  "ARIMA(1,1,0)" = c(1, 1, 0)
)

sarima_specs <- list(
  "SARIMA(1,1,1)(1,0,0)[4]" = list(order = c(1,1,1), seasonal = c(1,0,0)),
  "SARIMA(0,1,1)(1,0,0)[4]" = list(order = c(0,1,1), seasonal = c(1,0,0)),
  "SARIMA(1,1,1)(0,0,1)[4]" = list(order = c(1,1,1), seasonal = c(0,0,1)),
  "SARIMA(0,1,1)(0,0,1)[4]" = list(order = c(0,1,1), seasonal = c(0,0,1))
)

# Store training fits for plotting
arima_train_fits <- list()
for (nm in names(arima_specs)) {
  fit_obj <- forecast::Arima(hh_train, order = arima_specs[[nm]], include.mean = TRUE)
  arima_train_fits[[nm]] <- as.numeric(fitted(fit_obj))
}

sarima_train_fits <- list()
for (nm in names(sarima_specs)) {
  sp <- sarima_specs[[nm]]
  fit_obj <- forecast::Arima(hh_train, order = sp$order, seasonal = sp$seasonal,
                             include.mean = TRUE)
  sarima_train_fits[[nm]] <- as.numeric(fitted(fit_obj))
}

arima_train_df <- tibble(date = train_dates, Actual = as.numeric(hh_train))
for (nm in names(arima_train_fits)) {
  arima_train_df[[nm]] <- arima_train_fits[[nm]]
}
for (nm in names(sarima_train_fits)) {
  arima_train_df[[nm]] <- sarima_train_fits[[nm]]
}

arima_train_long <- arima_train_df |>
  pivot_longer(-date, names_to = "Model", values_to = "value")

ggplot(arima_train_long, aes(x = date, y = value, color = Model, linetype = Model)) +
  geom_line(linewidth = 0.8) +
  theme_minimal() +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Training set fit: ARIMA / SARIMA models (1977 Q1 -- 1996 Q4)",
    x = "Quarter", y = "Household net savings"
  )
```

```{r training RMSE comparison}
train_rmse <- function(actual, fitted) {
  valid <- !is.na(fitted)
  sqrt(mean((actual[valid] - fitted[valid])^2))
}

train_rmse_tbl <- tibble(
  Model = c("Persistence", "Average", "SES", "Holt Linear", "HW Additive",
            names(arima_train_fits), names(sarima_train_fits)),
  Train_RMSE = round(c(
    train_rmse(as.numeric(hh_train), persist_train_fitted),
    train_rmse(as.numeric(hh_train), avg_train_fitted),
    train_rmse(as.numeric(hh_train), ses_train_fitted),
    train_rmse(as.numeric(hh_train), holt_train_fitted),
    train_rmse(as.numeric(hh_train), hwa_train_fitted),
    sapply(arima_train_fits, \(f) train_rmse(as.numeric(hh_train), f)),
    sapply(sarima_train_fits, \(f) train_rmse(as.numeric(hh_train), f))
  ), 4)
) |>
  arrange(Train_RMSE)

# Print training RMSE table
train_rmse_tbl
```

## Benchmark Holdout Forecasts

We evaluate all benchmark rules on the 92-quarter holdout for reference.

```{r benchmarks}
ew_rmse_tbl <- tibble(Model = character(), Method = character(), RMSE = double())
ew_fc_store <- list()

# Persistence
res <- persist_fc(hh_train, hh_test)
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Persistence", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$Persistence <- res$fc

# Average
res <- iid_fc(hh_train, hh_test)
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Average", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$Average <- res$fc

# Average by Period
res <- iidbyperiod_fc(hh_train, hh_test, periodicity = 4)
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Average by Period", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$`Average by Period` <- res$fc

# SES
res <- esm_fc(hh_train, hh_test,
              alpha = hw_ses$alpha,
              level = as.numeric(hw_ses$coefficients[1]))
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "SES", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$SES <- res$fc

# Holt Linear
res <- lholt_fc(hh_train, hh_test,
                alpha = hw_holt$alpha,
                beta  = hw_holt$beta,
                level = as.numeric(hw_holt$coefficients[1]),
                slope = as.numeric(hw_holt$coefficients[2]))
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "Holt Linear", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$`Holt Linear` <- res$fc

# HW Additive
res <- aseason_fc(hh_train, hh_test,
                  alpha  = hw_add$alpha,
                  beta   = hw_add$beta,
                  gamma  = hw_add$gamma,
                  level  = as.numeric(hw_add$coefficients[1]),
                  slope  = as.numeric(hw_add$coefficients[2]),
                  season = as.numeric(hw_add$coefficients[3:6]))
ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = "HW Additive", Method = "Benchmark", RMSE = res$rmse)
ew_fc_store$`HW Additive` <- res$fc

ew_rmse_tbl
```

## Fixed-Parameter One-Step Forecasts (ARIMA / SARIMA)

For reference, we run fixed-parameter one-step-ahead forecasts on this longer holdout.

```{r ARIMA fixed parameter}
arima_fixed_results <- list()

for (nm in names(arima_specs)) {
  res <- arima_one_step_fc(hh_train, hh_test,
                           order = arima_specs[[nm]],
                           include.mean = TRUE)
  arima_fixed_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Fixed", RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (fixed)")]] <- res$fc
}

arima_fixed_tbl <- tibble(
  Model = names(arima_specs),
  AIC = sapply(arima_fixed_results, \(r) round(r$fit$aic, 2)),
  RMSE_Fixed = sapply(arima_fixed_results, \(r) round(r$rmse, 4))
)
arima_fixed_tbl
```

```{r SARIMA fixed parameter}
sarima_fixed_results <- list()
for (nm in names(sarima_specs)) {
  sp <- sarima_specs[[nm]]
  res <- sarima_one_step_fc(hh_train, hh_test,
                            order = sp$order,
                            seasonal = sp$seasonal,
                            include.mean = TRUE)
  sarima_fixed_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Fixed", RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (fixed)")]] <- res$fc
}

sarima_fixed_tbl <- tibble(
  Model = names(sarima_specs),
  AIC = sapply(sarima_fixed_results, \(r) round(r$fit$aic, 2)),
  RMSE_Fixed = sapply(sarima_fixed_results, \(r) round(r$rmse, 4))
)
sarima_fixed_tbl
```

## Rolling Window Forecasts

Now we re-estimate each model at every holdout step.

```{r ARIMA rolling window}
arima_ew_results <- list()
for (nm in names(arima_specs)) {
  res <- arima_expanding_window_fc(hh_train, hh_test,
                                   order = arima_specs[[nm]],
                                   include.mean = TRUE,
                                   window_size = R_window)
  arima_ew_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Rolling", 
                         RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (rolling)")]] <- res$fc
}
```

```{r SARIMA rolling window}
sarima_ew_results <- list()
for (nm in names(sarima_specs)) {
  sp <- sarima_specs[[nm]]
  res <- sarima_expanding_window_fc(hh_train, hh_test,
                                    order = sp$order,
                                    seasonal = sp$seasonal,
                                    include.mean = TRUE,
                                    window_size = R_window)
  sarima_ew_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Rolling", 
                         RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (rolling)")]] <- res$fc
}
```

## ARIMAX Rolling Window Forecast

### Variable Selection and Feasibility

We use the same CCF-justified regressors as in notebook 02: lagged log-differences
of GPDI and CPI (`ldgpdi_lag1`, `ldcpi_lag2`, `ldgpdi_lag3`). These were selected
based on significant cross-correlations at negative lags (the regressor leads
household savings).

**Feasibility of variable selection in the rolling window:**

- The CCF-based selection was originally performed on the notebook 02 training
  sample. For the rolling window we apply the same regressor set starting from
  the 1977--1996 initial window. This is conservative: if the leading
  relationship holds over the longer sample, it should manifest in sub-windows
  as well.
- Lagged regressors (`lag 1`, `lag 2`, `lag 3`) are available at forecast time
  by construction, so there is no look-ahead bias.
- The rolling window lets us empirically verify whether the regressor--response
  relationship persists as the training window rolls through different regimes.

```{r ARIMAX preprocessing}
# Build ARIMAX regressors from the full pre-2020 dataset
arimax_ew_df <- full_df |>
  arrange(date) |>
  mutate(
    ldgpdi      = log(gpdi) - log(lag(gpdi)),
    ldcpi       = log1p(cpi) - log1p(lag(cpi)),
    ldcpi_lag2  = lag(ldcpi, 2),
    ldgpdi_lag1 = lag(ldgpdi, 1),
    ldgpdi_lag3 = lag(ldgpdi, 3)
  ) |>
  drop_na(ldgpdi_lag1, ldcpi_lag2, ldgpdi_lag3)

arimax_vars <- c("ldgpdi_lag1", "ldcpi_lag2", "ldgpdi_lag3")
arimax_reduced_vars <- c("ldgpdi_lag1")

# Split at the initial training boundary — full regressors
x_ew_train <- arimax_ew_df |> filter(date <= initial_end) |>
  select(all_of(arimax_vars)) |> as.matrix()
x_ew_test <- arimax_ew_df |> filter(date > initial_end) |>
  select(all_of(arimax_vars)) |> as.matrix()

# Reduced regressors (only ldgpdi_lag1)
x_ew_train_r <- arimax_ew_df |> filter(date <= initial_end) |>
  select(all_of(arimax_reduced_vars)) |> as.matrix()
x_ew_test_r <- arimax_ew_df |> filter(date > initial_end) |>
  select(all_of(arimax_reduced_vars)) |> as.matrix()

# Align hh_save to the ARIMAX-available dates (loses ~4 initial obs due to lagging)
hh_train_x <- ts(
  arimax_ew_df |> filter(date <= initial_end) |> pull(hh_save),
  start = c(year(min(arimax_ew_df$date)), quarter(min(arimax_ew_df$date))),
  frequency = 4
)
hh_test_x <- ts(
  arimax_ew_df |> filter(date > initial_end) |> pull(hh_save),
  start = c(
    year(min(arimax_ew_df |> filter(date > initial_end) |> pull(date))),
    quarter(min(arimax_ew_df |> filter(date > initial_end) |> pull(date)))
  ),
  frequency = 4
)

cat("ARIMAX initial training window:", length(hh_train_x))
cat("ARIMAX holdout window:", length(hh_test_x))
```

### ARIMAX Coefficient Significance (Initial Window)

We fit ARIMAX(1,1,1) on the initial 1977--1996 window to inspect whether the
exogenous coefficients are significant before committing to the full rolling
window evaluation.

```{r ARIMAX initial fit}
arimax_specs <- list(
  "ARIMAX(1,1,1) (Full)" = c(1, 1, 1),
  "ARIMAX(0,1,1) (Full)" = c(0, 1, 1),
  "ARIMAX(1,1,2) (Full)" = c(1, 1, 2),
  "ARIMAX(2,1,2) (Full)" = c(2, 1, 2)
)

arimax_reduced_specs <- list(
  "ARIMAX(1,1,1) (Reduced)" = c(1, 1, 1),
  "ARIMAX(0,1,1) (Reduced)" = c(0, 1, 1),
  "ARIMAX(1,1,2) (Reduced)" = c(1, 1, 2),
  "ARIMAX(2,1,2) (Reduced)" = c(2, 1, 2)
)

fit_arimax_init <- forecast::Arima(hh_train_x, order = c(1, 1, 1),
                                    xreg = x_ew_train, include.mean = TRUE)
summary(fit_arimax_init)

# Extract coefficient significance
coefs <- coef(fit_arimax_init)
ses   <- sqrt(diag(vcov(fit_arimax_init)))
z_vals <- coefs / ses
p_vals <- 2 * (1 - pnorm(abs(z_vals)))

coef_tbl <- tibble(
  Coefficient = names(coefs),
  Estimate = round(coefs, 6),
  SE = round(ses, 6),
  z = round(z_vals, 3),
  p_value = round(p_vals, 4),
  Significant = ifelse(p_vals < 0.05, "Yes", "No")
)
coef_tbl
```

### ARIMAX Fixed-Parameter Forecasts (Full Regressors)

```{r ARIMAX fixed parameter full}
arimax_fixed_results <- list()
for (nm in names(arimax_specs)) {
  res <- arimax_one_step_fc(hh_train_x, hh_test_x,
                            xreg_train = x_ew_train,
                            xreg_holdout = x_ew_test,
                            order = arimax_specs[[nm]],
                            include.mean = TRUE)
  arimax_fixed_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Fixed", RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (fixed)")]] <- res$fc
}

arimax_fixed_tbl <- tibble(
  Model = names(arimax_specs),
  AIC = sapply(arimax_fixed_results, \(r) round(r$fit$aic, 2)),
  RMSE_Fixed = sapply(arimax_fixed_results, \(r) round(r$rmse, 4))
)
arimax_fixed_tbl
```

### ARIMAX Rolling Window Forecasts (Full Regressors)

```{r ARIMAX rolling window full}
arimax_ew_results <- list()
for (nm in names(arimax_specs)) {
  res <- arimax_expanding_window_fc(hh_train_x, hh_test_x,
                                     xreg_train = x_ew_train,
                                     xreg_holdout = x_ew_test,
                                     order = arimax_specs[[nm]],
                                     include.mean = TRUE,
                                     window_size = R_window)
  arimax_ew_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Rolling",
                         RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (rolling)")]] <- res$fc
}
```

### ARIMAX Fixed-Parameter Forecasts (Reduced: `ldgpdi_lag1` only)

```{r ARIMAX fixed parameter reduced}

# Re-run fixed-parameter forecasts with only the significant regressor (ldgpdi_lag1)
arimax_r_fixed_results <- list()
for (nm in names(arimax_reduced_specs)) {
  res <- arimax_one_step_fc(hh_train_x, hh_test_x,
                            xreg_train = x_ew_train_r,
                            xreg_holdout = x_ew_test_r,
                            order = arimax_reduced_specs[[nm]],
                            include.mean = TRUE)
  arimax_r_fixed_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Fixed", RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (fixed)")]] <- res$fc
}

arimax_r_fixed_tbl <- tibble(
  Model = names(arimax_reduced_specs),
  AIC = sapply(arimax_r_fixed_results, \(r) round(r$fit$aic, 2)),
  RMSE_Fixed = sapply(arimax_r_fixed_results, \(r) round(r$rmse, 4))
)
arimax_r_fixed_tbl
```

### ARIMAX Rolling Window Forecasts (Reduced: `ldgpdi_lag1` only)

```{r ARIMAX rolling window reduced}

arimax_r_ew_results <- list()
for (nm in names(arimax_reduced_specs)) {
  res <- arimax_expanding_window_fc(hh_train_x, hh_test_x,
                                     xreg_train = x_ew_train_r,
                                     xreg_holdout = x_ew_test_r,
                                     order = arimax_reduced_specs[[nm]],
                                     include.mean = TRUE,
                                     window_size = R_window)
  arimax_r_ew_results[[nm]] <- res
  ew_rmse_tbl <- add_row(ew_rmse_tbl, Model = nm, Method = "Rolling",
                         RMSE = res$rmse)
  ew_fc_store[[paste0(nm, " (rolling)")]] <- res$fc
}
```

## Comparison Tables

### Fixed vs Rolling Window

```{r comparison table}

# ARIMA: fixed vs rolling
arima_compare <- tibble(
  Model = names(arima_specs),
  RMSE_Fixed = sapply(arima_fixed_results, \(r) round(r$rmse, 4)),
  RMSE_Rolling = sapply(arima_ew_results, \(r) round(r$rmse, 4)),
  Improvement_pct = round((RMSE_Fixed - RMSE_Rolling) / RMSE_Fixed * 100, 2)
)

arima_compare
```

```{r}
# SARIMA: fixed vs rolling
sarima_compare <- tibble(
  Model = names(sarima_specs),
  RMSE_Fixed = sapply(sarima_fixed_results, \(r) round(r$rmse, 4)),
  RMSE_Rolling = sapply(sarima_ew_results, \(r) round(r$rmse, 4)),
  Improvement_pct = round((RMSE_Fixed - RMSE_Rolling) / RMSE_Fixed * 100, 2)
)

sarima_compare
```

```{r}
# ARIMAX (Full): fixed vs rolling
arimax_compare <- tibble(
  Model = names(arimax_specs),
  RMSE_Fixed = sapply(arimax_fixed_results, \(r) round(r$rmse, 4)),
  RMSE_Rolling = sapply(arimax_ew_results, \(r) round(r$rmse, 4)),
  Improvement_pct = round((RMSE_Fixed - RMSE_Rolling) / RMSE_Fixed * 100, 2)
)
arimax_compare
```

```{r}
# ARIMAX (Reduced): fixed vs rolling
arimax_r_compare <- tibble(
  Model = names(arimax_reduced_specs),
  RMSE_Fixed = sapply(arimax_r_fixed_results, \(r) round(r$rmse, 4)),
  RMSE_Rolling = sapply(arimax_r_ew_results, \(r) round(r$rmse, 4)),
  Improvement_pct = round((RMSE_Fixed - RMSE_Rolling) / RMSE_Fixed * 100, 2)
)
arimax_r_compare
```

### Full RMSE Ranking (Benchmarks + ARIMA/SARIMA/ARIMAX)

```{r overall RMSE ranking}
ew_rmse_tbl <- ew_rmse_tbl |>
  mutate(RMSE = round(RMSE, 4)) |>
  arrange(RMSE)

ew_rmse_tbl

best_ew_model <- ew_rmse_tbl |> slice(1)

best_ew_model
```

```{r persistence comparison}

# How do rolling window models compare to persistence?
persist_rmse <- ew_rmse_tbl |> 
  filter(Model == "Persistence") |> 
  pull(RMSE)


# Print models that beat persistence
ew_rmse_tbl |> 
  filter(RMSE < persist_rmse & Method == "Rolling")
```

```{r save rolling window results}
saveRDS(ew_rmse_tbl, file = "../output/ew_rmse_tbl.rds")
```

## Holdout Visualization

### Benchmarks vs Best Rolling Window Models

```{r all models holdout, fig.width=12, fig.height=10}

best_arima_nm  <- arima_compare$Model[which.min(arima_compare$RMSE_Rolling)]
best_sarima_nm <- sarima_compare$Model[which.min(sarima_compare$RMSE_Rolling)]
arimax_all_compare <- bind_rows(arimax_compare, arimax_r_compare)
best_arimax_nm <- arimax_all_compare$Model[which.min(arimax_all_compare$RMSE_Rolling)]

fc_all_df <- tibble(
  date = ew_test_df$date,
  Actual = as.numeric(hh_test),
  Persistence = as.numeric(ew_fc_store$Persistence),
  Average = as.numeric(ew_fc_store$Average),
  SES = as.numeric(ew_fc_store$SES),
  `Holt Linear` = as.numeric(ew_fc_store$`Holt Linear`),
  `HW Additive` = as.numeric(ew_fc_store$`HW Additive`)
)
fc_all_df[[paste0(best_arima_nm, " (roll)")]] <-
  as.numeric(ew_fc_store[[paste0(best_arima_nm, " (rolling)")]])
fc_all_df[[paste0(best_sarima_nm, " (roll)")]] <-
  as.numeric(ew_fc_store[[paste0(best_sarima_nm, " (rolling)")]])
fc_all_df[[paste0(best_arimax_nm, " (roll)")]] <-
  as.numeric(ew_fc_store[[paste0(best_arimax_nm, " (rolling)")]])

fc_all_long <- fc_all_df |>
  pivot_longer(-date, names_to = "Model", values_to = "value")

# Faceted panel: each model vs Actual (grey)
actual_df <- fc_all_long |> filter(Model == "Actual")
model_df  <- fc_all_long |> filter(Model != "Actual")

ggplot(model_df, aes(x = date)) +
  geom_line(data = actual_df |> select(-Model) |>
              crossing(Model = unique(model_df$Model)),
            aes(y = value), color = "grey60", linewidth = 0.5) +
  geom_line(aes(y = value), color = "#1b9e77", linewidth = 0.7) +
  facet_wrap(~ Model, ncol = 2) +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold")) +
  labs(
    title = "Holdout forecasts vs Actual (grey): Benchmarks & best rolling window models",
    x = "Quarter", y = "Household net savings"
  )
```

### Fixed vs Rolling: Best ARIMA, SARIMA, ARIMAX

```{r fixed vs rolling comparison, fig.width=12, fig.height=8}
# Build a combined faceted comparison: Fixed vs Rolling for best of each family
compare_models <- list(
  ARIMA  = best_arima_nm,
  SARIMA = best_sarima_nm,
  ARIMAX = best_arimax_nm
)

compare_df <- tibble(date = ew_test_df$date, Actual = as.numeric(hh_test))
compare_long_list <- list()

for (family in names(compare_models)) {
  nm <- compare_models[[family]]
  panel_df <- tibble(
    date = ew_test_df$date,
    Actual        = as.numeric(hh_test),
    `Fixed params` = as.numeric(ew_fc_store[[paste0(nm, " (fixed)")]]),
    `Rolling window` = as.numeric(ew_fc_store[[paste0(nm, " (rolling)")]]),
    Persistence   = as.numeric(ew_fc_store$Persistence)
  ) |>
    pivot_longer(-date, names_to = "Method", values_to = "value") |>
    mutate(Panel = paste0(family, ": ", nm))
  compare_long_list[[family]] <- panel_df
}

compare_long <- bind_rows(compare_long_list)

ggplot(compare_long, aes(x = date, y = value, color = Method, linetype = Method)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ Panel, ncol = 1, scales = "free_y") +
  theme_minimal(base_size = 11) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom") +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Fixed vs Rolling Window: Best ARIMA, SARIMA, ARIMAX",
    x = "Quarter", y = "Household net savings"
  )
```

```{r Scatter matrix of forecasts}
# Scatter matrix of forecasts: Persistence vs HW vs best ARIMA/SARIMA/ARIMAX
scatter_df <- tibble(
  Persistence = as.numeric(ew_fc_store$Persistence),
  SES = as.numeric(ew_fc_store$`SES`),
  ARIMA = as.numeric(ew_fc_store[[paste0(best_arima_nm, " (rolling)")]]),
  SARIMA = as.numeric(ew_fc_store[[paste0(best_sarima_nm, " (rolling)")]]),
  ARIMAX = as.numeric(ew_fc_store[[paste0(best_arimax_nm, " (rolling)")]])
)

pairs(scatter_df, main = "Scatter matrix: Persistence vs best ARIMA/SARIMA/ARIMAX",
      pch = 19, col = "blue", cex = 0.6)
```

## Interpretation

- The rolling window (R = 80 quarters) re-estimates model parameters at each step using only the most recent 20 years of data. This prevents observations from old regimes from diluting current parameter estimates, and better tracks structural shifts over the 92-quarter holdout.
- Comparing the rolling window ARIMA/SARIMA/ARIMAX models directly against benchmark rules (Persistence, Average, SES, Holt, HW Additive) reveals whether the added complexity of parametric models is justified.
- If the rolling window RMSE is similar to the fixed-parameter RMSE, this confirms that the series is close to a random walk, and parameter re-estimation adds little value.
- The persistence benchmark remains the key reference: any model must beat persistence to justify its complexity.

**ARIMAX observations:**

- We evaluate two regressor sets: **Full** (`ldgpdi_lag1`, `ldcpi_lag2`, `ldgpdi_lag3`) includes all CCF-justified lags; **Reduced** (`ldgpdi_lag1` only) retains only the regressor that is statistically significant at the 5% level in the initial window fit.
- If the Reduced ARIMAX matches or beats the Full ARIMAX on rolling-window RMSE, parameter estimation noise from the insignificant regressors is hurting forecasts and the parsimonious specification is preferred.
- The coefficient significance table from the initial window fit provides a first diagnostic: insignificant coefficients suggest the ARIMA residuals are already orthogonal to the regressors. In that case, including them inflates estimation variance and may worsen out-of-sample performance.
- The rolling window is particularly informative for evaluating regressor usefulness: if ARIMAX RMSE tracks closely with or exceeds ARIMA RMSE as the window rolls through different regimes, the exogenous variables should be dropped in favour of the more parsimonious ARIMA specification.
