---
title: "EDA"
output:
  pdf_document: default
  html_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(forecast)
library(knitr)
library(tseries)
```

### Load the Harmonized Data and a First Glimpse
```{r data loading, include=TRUE}
macro_df <- read_csv("../data/macro_panel_wide_raw.csv", show_col_types = FALSE) |>
  mutate(quarterly_date = as.Date(quarterly_date)) |>
  arrange(quarterly_date)

glimpse(macro_df)
summary(macro_df)
```

```{r checking missing}
na_summary <- macro_df |>
  summarise(across(-quarterly_date, ~ sum(is.na(.)))) |>
  pivot_longer(
    cols = everything(),
    names_to = "series",
    values_to = "n_missing"
  ) |>
  arrange(desc(n_missing))

na_summary

# Remove the NA values at the end
macro_df <- macro_df |>
  filter(quarterly_date <= as.Date("2023-12-31"))
```
```{r distribution of key variable values}
key_vars <- macro_df |>
  select(-quarterly_date) |>
  pivot_longer(
    cols = everything(),
    names_to = "series",
    values_to = "value"
  )

ggplot(key_vars, aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ series, scales = "free", ncol = 2)+ 
  theme_minimal() +
  ggtitle("Distribution of key macro series values") +
  xlab("Value") +
  ylab("Frequency")
```


```{r Plot all series}
macro_long <- macro_df |>
  pivot_longer(
    cols = -quarterly_date,
    names_to = "series",
    values_to = "value"
  )

ggplot(macro_long, aes(x = quarterly_date, y = value)) +
  geom_line() +
  facet_wrap(~ series, scales = "free_y", ncol = 2) +
  theme_minimal() +
  labs(
    title = "All macro series over time",
    x = "Quarter",
    y = "Value"
  )
```

```{r EDA dataset}
eda_df <- macro_df |>
  select(
    quarterly_date,
    national_net_savings,
    household_net_savings,
    corporate_net_savings,
    gross_domestic_income,
    household_consumption,
    unemployment_rate,
    cpi_inflation_indicator,
    market_interest_rates
  ) |>
  rename(
    date = quarterly_date,
    nat_save = national_net_savings,
    hh_save = household_net_savings,
    corp_save = corporate_net_savings,
    gdi = gross_domestic_income,
    cons = household_consumption,
    unemp = unemployment_rate,
    cpi = cpi_inflation_indicator,
    rate = market_interest_rates
  ) |>
  mutate(
    hh_rate = (hh_save / gdi) / 10000, 
    nat_rate = (nat_save / gdi) / 10000,
    corp_rate = (corp_save / gdi) / 10000
  ) |>
  arrange(date)

glimpse(eda_df)
summary(eda_df)

```
```{r Save the EDA dataset for later use}
saveRDS(eda_df, file = "../data/eda_df_nodiff.rds")
```


```{r Compare the three net savings series}
savings_long <- eda_df |>
  select(date, nat_save, hh_save, corp_save) |>
  pivot_longer(
    cols = -date,
    names_to = "series",
    values_to = "value"
  )

options(repr.plot.width = 10, repr.plot.height = 5)

ggplot(savings_long, aes(x = date, y = value, color = series)) +
  geom_line(linewidth = 0.9) +
  theme_minimal() +
  labs(
    title = "National, household, and corporate net savings",
    x = "Quarter",
    y = "Net savings",
    color = "Series"
  )
```

```{r Focus on household and corporate net savings}
main_response_long <- eda_df |>
  select(date, hh_save, corp_save) |>
  pivot_longer(
    cols = -date,
    names_to = "series",
    values_to = "value"
  )

ggplot(main_response_long, aes(x = date, y = value)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ series, scales = "free_y", ncol = 1) +
  theme_minimal() +
  labs(
    title = "Main response series",
    x = "Quarter",
    y = "Net savings"
  )
```

### Transformations and Stationarity Checks

```{r Focused GDI EDA}
eda_df <- eda_df |>
  mutate(
    gdi_qoq = 100 * (gdi / lag(gdi) - 1),
    gdi_yoy = 100 * (gdi / lag(gdi, 4) - 1),
    ldgdi = log(gdi) - log(lag(gdi))
  )

summary(select(eda_df, gdi, gdi_qoq, gdi_yoy, ldgdi))
```

```{r GDI EDA}
p1 <- ggplot(eda_df, aes(x = date, y = gdi)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Gross domestic income",
    x = "Quarter",
    y = "GDI"
  )

p2 <- ggplot(eda_df, aes(x = date, y = gdi_yoy)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "GDI year-over-year change (%)",
    x = "Quarter",
    y = "YoY % change"
  )

p3 <- ggplot(eda_df, aes(x = date, y = ldgdi*100)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Log GDI change (%)",
    x = "Quarter",
    y = "YoY % change (log)"
  )

p1
p2
p3
```
We will apply the Inverse Hyperbolic Sine (IHS) transformation to the main response series, which is similar to log transformation but can handle zero and negative values. This will help stabilize the variance and make the series more stationary for modeling purposes. We will also create differenced versions of the series to check for stationarity and lead-lag relationships with other variables.

```{r Simple transformations for the main outcome series}

eda_df <- eda_df |>
  mutate(
    dhh_rate = hh_rate - lag(hh_rate),
    dcorp_rate = corp_rate - lag(corp_rate),
    dnat_rate = nat_rate - lag(nat_rate),
    unemp_diff = unemp - lag(unemp),
    rate_diff = rate - lag(rate),
    ldcons = log1p(cons) - log1p(lag(cons)),
    ldcpi = log1p(cpi) - log1p(lag(cpi))
  )

summary(select(
  eda_df,
  dhh_rate,
  dcorp_rate,
  dnat_rate,
  unemp_diff,
  rate_diff,
  ldcons,
  ldcpi
))
```

```{r Plot transformed household and corporate series}
transformed_long <- eda_df |>
  select(date, dhh_rate, dcorp_rate) |>
  pivot_longer(
    cols = -date,
    names_to = "series",
    values_to = "value"
  )

ggplot(transformed_long, aes(x = date, y = value*100)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ series, scales = "free_y", ncol = 1) +
  theme_minimal() +
  labs(
    title = "Percentage change of household and corporate net saving Rates",
    x = "Quarter",
    y = "Percentage change (%) (By log)"
  )
```

```{r Time-series diagnostics for household net savings}
hh_ts <- ts(
  eda_df$hh_rate,
  start = c(year(min(eda_df$date)), quarter(min(eda_df$date))),
  frequency = 4
)

dhh_ts <- diff(hh_ts)

par(mfrow = c(2, 2))
plot(hh_ts, main = "Household net savings")
plot(dhh_ts, main = "First difference: household net savings")
acf(na.omit(dhh_ts), lag.max = 20, main = "ACF: diff household savings")
pacf(na.omit(dhh_ts), lag.max = 20, main = "PACF: diff household savings")
par(mfrow = c(1, 1))

adf.test(na.omit(dhh_ts))
```

```{r Time-series diagnostics for corporate net savings}
corp_ts <- ts(
  eda_df$corp_rate,
  start = c(year(min(eda_df$date)), quarter(min(eda_df$date))),
  frequency = 4
)

dcorp_ts <- diff(corp_ts)

par(mfrow = c(2, 2))
plot(corp_ts, main = "Corporate net savings")
plot(dcorp_ts, main = "First difference: corporate net savings")
acf(na.omit(dcorp_ts), lag.max = 20, main = "ACF: diff corporate savings")
pacf(na.omit(dcorp_ts), lag.max = 20, main = "PACF: diff corporate savings")
par(mfrow = c(1, 1))

adf.test(na.omit(dcorp_ts))
```


```{r Benchmark diagnostics for national net savings}
nat_ts <- ts(
  eda_df$nat_rate,
  start = c(year(min(eda_df$date)), quarter(min(eda_df$date))),
  frequency = 4
)

dnat_ts <- diff(nat_ts)

par(mfrow = c(2, 2))
plot(nat_ts, main = "National net savings")
plot(dnat_ts, main = "First difference: national net savings")
acf(na.omit(dnat_ts), lag.max = 20, main = "ACF: diff national savings")
pacf(na.omit(dnat_ts), lag.max = 20, main = "PACF: diff national savings")
par(mfrow = c(1, 1))

adf.test(na.omit(dnat_ts))
```


```{r Lead-lag checks with GDI, unemployment, and interest rates}
par(mfrow = c(2, 3))

ccf(na.omit(eda_df$ldgdi), na.omit(eda_df$dhh_rate),
    main = "log GDI YoY vs household savings")
ccf(na.omit(eda_df$ldgdi), na.omit(eda_df$dcorp_rate),
    main = "log GDI YoY vs corporate savings")

ccf(na.omit(eda_df$unemp_diff), na.omit(eda_df$dhh_rate),
    main = "Unemployment vs household savings")
ccf(na.omit(eda_df$unemp_diff), na.omit(eda_df$dcorp_rate),
    main = "Unemployment vs corporate savings")

ccf(na.omit(eda_df$ldcons), na.omit(eda_df$dhh_rate),
    main = "Consumption vs household savings")
ccf(na.omit(eda_df$ldcons), na.omit(eda_df$dcorp_rate),
    main = "Consumption vs corporate savings")

par(mfrow = c(1, 1))
```
From the above, we see the following:

- 

```{r Simple scatterplot matrix}
pairs(
  eda_df |>
    select(hh_rate, corp_rate, gdi, cons, unemp, cpi, rate) |>
    na.omit(),
  main = "Scatterplot matrix for key variables"
)
```
